---
title: "Aggregate GeoTIFF"
author: "Beni Stocker and Ting Tan"
date: "2025-02-19"
output:
  html_document:
    fig_align: 'center'
bibliography: literatures.bib
csl: apa.csl
---

```{r setup, include=FALSE}
library(terra)
```
## 📊 Aggregation

The **very high-resolution global canopy height dataset** ([@langHighresolutionCanopyHeight2023](https://www.nature.com/articles/s41559-023-02206-6)) is available for download from [ETH Zurich](https://doi.org/10.3929/ethz-b-000609802). In contrast, the **global Topographic Wetness Index (TWI)** dataset ([@marthewsHighresolutionGlobalTopographic2015](https://hess.copernicus.org/articles/19/91/2015/)) has a coarser spatial resolution of 15 arcseconds (approximately 450 meters at the equator) and is accessible via the [CEH Catalogue](https://catalogue.ceh.ac.uk/documents/6b0c4358-2bf3-4924-aa8f-793d468b92be).

To ensure spatial compatibility for subsequent analysis, the canopy height dataset must be **aggregated** to match the 15 arcsecond resolution of the TWI data.

## 🧩 Combination

After aggregating the global canopy height data to match the 15 arcsecond (~450 m) resolution of the TWI dataset, the next step is to **combine multiple raster tiles into a single mosaic** to facilitate global-scale analysis.

This step is implemented in the script:  
📄 `analysis/01_aggregate_geotiff.R`

```{r Visualization, eval = TRUE}

# Read the mosaic file
file_vegh_450m_mosaic <- "/data_2/scratch/ting/data/vegh_450m/vegh_450m_2020_mosaic.nc"
# Plot the mosaic raster after writeCDF
plot(terra::rast(file_vegh_450m_mosaic))
```

## 📚 References
```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here("vignettes/01_aggregate_geotiff.Rmd"))
```
