---
title: "Aggregate GeoTIFF"
author: "Beni Stocker and Ting Tan"
date: "2025-02-19"
output: html_document
---

```{r setup, include=FALSE}
library(terra)
```

# Aggregation
The very high-resolution global canopy height dataset ([Lang et al., 2023](https://www.nature.com/articles/s41559-023-02206-6)) is available for download from [ETH Zurich](https://doi.org/10.3929/ethz-b-000609802). In contrast, the global Topographic Wetness Index (TWI) dataset ([Marthew, 2015](https://hess.copernicus.org/articles/19/91/2015/)) has a coarser resolution of 15 arcseconds (~450m at the Equator) and is available from the [CEH Catalogue](https://catalogue.ceh.ac.uk/documents/6b0c4358-2bf3-4924-aa8f-793d468b92be). To ensure compatibility for further analysis, the global canopy height data must first be aggregated to match the 15 arcsecond resolution of the TWI dataset.

This step finish by "analysis/01_aggregate_geotiff.R"

# Combination
After aggregating the global canopy height data to match the 15 arcsecond (~450m) resolution of the TWI dataset, the next step is to combine multiple raster tiles into a single mosaic for better 

This step finish by "analysis/01_aggregate_geotiff.R"

```{r Visualization, eval = TRUE}

# Read the mosaic file
file_vegh_mosaic <- file.path(dirname(dir_vegh_450m), "vegh_450m_2020_mosaic.nc")

# Plot the mosaic raster after writeCDF
plot(terra::rast(file_vegh_mosaic))
```

