---
title: "Split Window Correlation"
author: "Ting Tan"
date: "2025-03-27"
output: html_document
---

# Introduction
This script tests the split-window correlation method without applying a land cover mask to exclude cropland and urban areas. Instead, a small random region in Africa (0, 1, 5.5, 6.5) is selected for analysis. The goal is to evaluate the correlation between the Topographic Wetness Index (TWI) and vegetation height (Vegh) within this region.

## Required Libraries and Dependencies
To execute this script, several R libraries are loaded, including `terra`, `sf`, `dplyr`, and `ggplot2`. Additionally, a custom function from the `split_window_correlation.R` script is sourced.

```{r setup, include=FALSE}
library(terra)
library(sf)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(ggpointdensity)
library(ncdf4)
library(cowplot)
library(scico)
library(ggmap)
source(here::here("R/split_window_correlation.R"))
```

## Configuration
The paths to the necessary input raster files and the output correlation analysis file are defined.

```{r Configuration, include=TRUE, eval=TRUE}
file_vegh_450m_mosaic <- "/data_2/scratch/ting/data/vegh_450m/vegh_450m_2020_mosaic.nc"
file_ga2 <- "/data/archive/gti_marthews_2015/data/ga2.nc"  # Target raster file path
output_dir <- "/data_2/scratch/ting/data_temp"
file_correlation_nc <- "/data_2/scratch/ting/data/correlation_analysis.nc"  # Output file
```

## Objective
The objective of this analysis is to crop a specific area and test the split-window correlation method. The correlation between the TWI and vegetation height will be examined using spatial windows.

## Splitting Window Analysis

### Step 1: Preprocess
The selected region is cropped from the input raster datasets.

```{r splitting_window_analysis, include=TRUE, eval=TRUE}

ext_test <- list("test_region_af" = ext(0, 1, 5.5, 6.5))
region_name <- names(ext_test)[1]
  
merged_file <- preprocess_all_regions(
  ext_list = ext_test,
  twi_file = file_ga2,
  vegh_file = file_vegh_450m_mosaic,
  output_dir = output_dir
)

rm_test <- rast(merged_file[[1]])
names(rm_test) <- c("twi", "vegh")

plot(rm_test)
```

### Step 2: data analysis

* Spatial windows are generated for the merged raster dataset, allowing for localized correlation analysis. The dataset is checked for missing values.
* Correlations between TWI and vegetation height are computed for each spatial window.

```{r}
windowed_data <- create_spatial_windows(rm_test, 12)

colSums(is.na(windowed_data))
str(windowed_data)
colnames(windowed_data)

correlation_df <- calculate_window_correlations(windowed_data)
summary(correlation_df)
```

### Step 3: Visualization of Data
```{r}
plot1 <- plot_twi(windowed_data)
plot2 <- plot_vegh(windowed_data)
plot3 <- plot_window_distribution(windowed_data)
plot4 <- plot_correlation_vs_pixel_count(correlation_df)
plot5 <- plot_corr(correlation_df)

# plot6 <- ggplot() + theme_void()
plot6 <- plot_img(ext_test[[1]])

```

### Final Visualization
The generated plots are arranged and displayed for comparative analysis.
```{r, fig.width=12, fig.height=8}
all_plots <- list(plot1, plot2, plot3, plot4, plot5, plot6)

combined_plot <- plot_grid(
  plotlist = all_plots,
  ncol = 3,
  align = "hv"
) + theme(plot.background = element_rect(fill = "white", color = "white"))

# Save output plot
output_file <- here(
  "data",
  "figures",
  paste0("02_2_combined_plot_", region_name, ".png")
)

ggsave(
  output_file,
  combined_plot,
  width = 20,
  height = 10,
  dpi = 300,
  bg = "white"
)

knitr::include_graphics(output_file)
```

## Conclusion
This analysis successfully demonstrates the split-window correlation approach for a small test region in Africa. The results help understand the spatial relationship between TWI and vegetation height, providing valuable insights for further research and broader-scale applications.

```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here("vignettes/02_1_splitting_window_correlation_test.Rmd"))
```
