---
title: "Split Window Correlation -- validation the patterns under the framework of Fan et al. "
author: "Ting Tan"
date: "2025-04-02"
output:
  html_document:
    fig_align: 'center'
---

```{r setup, include=FALSE, eval = TRUE}

knitr::opts_chunk$set(
  comment = "#>", echo = FALSE, fig.width = 10, message = FALSE
)

library(terra)
library(sf)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(scico)
library(ggpointdensity)
library(ncdf4)
library(cowplot)
library(furrr)
library(knitr)
library(ggmap)
library(segmented)
library(progressr)
library(cli)
handlers("cli")

source(here::here("R/split_window_correlation.R"))
```
# Introduction

We set up 5 regions to validate the vegetation patterns under different hyrological conditions between valley and hilltop [@fanHillslopeHydrologyGlobal2019], spanning climatic gradients from hot to dry, cod to wet, and high to low relief.

## Objective

This document provides the step-by-step process for cropping, merging, performing correlation analysis, and visualizing the results using TWI (Topographic Wetness Index) and Tree Height data.

```{r Configuration}
file_vegh_450m_mosaic <- "/data_2/scratch/ting/data/vegh_450m/vegh_450m_2020_mosaic.nc"
file_ga2 <- "/data/archive/gti_marthews_2015/data/ga2.nc"  # Target raster file path
file_correlation_nc <- "/data_2/scratch/ting/data/correlation_analysis.nc"  # Output file
temp_dir <- "/data_2/scratch/ting/data_temp"
file_modis_landcover <- "/data/scratch/bstocker/landcover/modis_landcover__LPDAAC__v5.1__0.05deg__2010.nc"
file_modis_landcover1 <- "/data/archive/landcovermodis_friedl_2015/data/MODIS-C006_MOD15A2_LAI_FPAR_zmaw/MODIS-C006_MOD15A2__LAI_FPAR__LPDAAC__GLOBAL_0.5degree__UHAM-ICDC__2000_2018__MON__fv0.02.nc"

# 加载 NetCDF 文件
# modis_raster <- rast(file_modis_landcover1)
# nlyr(modis_raster)
# layer_names <- names(modis_raster)
# layer_names[3000:4000]
# print(modis_raster[[232]])
```

## Implementation

### 1. Crop the TWI and Tree Height Layers

**Input:** TWI and Tree Height rasters.  
**Output:** Cropped rasters for each region using the specified extents.

### 2. Merge the Two Layers (TWI and Tree Height)

**Input:** Cropped TWI and Tree Height rasters.  
**Output:** Merged raster with both TWI and Tree Height data.

### 3. Create 12x12 Windows with 0.05° Resolution

**Input:** The merged raster for each region.  
**Output:** 12x12 spatial windows for correlation analysis.

---

### 4. Pearson Correlation Analysis within Each Window

**Input:** 12x12 spatial windows.  
**Output:** Correlation matrix (Pearson r) for TWI and Tree Height.

### 5. Visualization

**Input:** TWI, Tree Height, Pearson Correlation (r), satellite images, MODIS land cover, and scatter plots for 3 random windows.  
**Output:** Combined visual output showing correlations, rasters, and MODIS land cover.

```{r DataPre}

ext_valids <- list(
  "desert_riparian" = ext(-112, -109, 28, 31),
  "mediterranean" = ext(-122.5, -119.5, 35, 38),
  "forest_savanna" = ext(28, 30, -6.5, -4.5),
  "waterlogged_pantanal" = ext(-58.5, -56, -19, -16),
  "white_sand" = ext(-68.5, -66.5, 1, 3)
)

# Prepare the dataset by combining both TWI and VEGH layers.
# The combined raster is saved to disk, and the file path is returned.
# This avoids passing large raster objects directly into parallel processing!!!
# Note: The processed raster files are not deleted automatically—
#       please delete them manually if no longer needed.
output_file <-here::here("data/preprocessed_info_validations.csv")
prep_info_valid <- generate_prep_tiles_info(
   ext_valids, file_ga2, file_vegh_450m_mosaic, temp_dir,
   output_file = output_file)

```

```{r parallel}
prep_info_valid <- readr::read_csv(info_file)
# Clear memory and set up parallel processing
gc()
plan(multisession, workers = 1)

log_file <- here::here("data/figures/processing_log.txt")
writeLines("=== Region Processing Log ===\n", log_file)

# Begin parallel processing with progress bar
with_progress({
  pb <- progressor(along = region_names)

  results <- future_pmap(
    prep_tiles_info,
    function(name, xmin, xmax, ymin, ymax, file){
      pb(sprintf("Processing %s", name))

      # Log inputs for debugging
      message("Processing region: ", name)
      message("raster_file for ", name, ": ", file)
      message("region_ext for ", name, ": ", xmin, xmax, ymin, ymax)

      # Load the raster file
      merged_raster <- suppressWarnings(terra::rast(file))
      names(merged_raster) <- c("twi", "vegh")

      # Process and analyze
      windowed_data <- create_spatial_windows(merged_raster)
      # correlation_df <- calculate_window_correlations1(windowed_data)
      correlation_df <- calculate_window_correlations2(windowed_data)

      # Debug output: NA count
      message(name, " NA count:")
      print(colSums(is.na(correlation_df)))

      # Generate plots
      plot1 <- plot_twi(windowed_data, xmin, xmax, ymin, ymax)
      plot2 <- plot_vegh(windowed_data, xmin, xmax, ymin, ymax)
      plot3 <- plot_corr(correlation_df, xmin, xmax, ymin, ymax)
      plot4 <- plot_correlation_vs_pixel_count(correlation_df)
      # plot5 <- plot_img(region_ext)
      
      # use the calculate_window_correlations2
      plot5 <- plot_peak(correlation_df)
      
      plot6 <- plot_landcover(file_modis_landcover, ext(xmin, xmax, ymin, ymax))
      
      # use the calculate_window_correlations1
      plots <- plot_random_windows(correlation_df, seed = 123)
      plot9 <- plot_overview(windowed_data)
      
      all_plots <- c(list(plot1, plot2, plot3, plot4, plot5, plot6),
                     plots,
                     list(plot9))
      
      combined_plot <- plot_grid(plotlist = all_plots, ncol = 3,
                                 align = "hv", axis = "tblr") + 
        theme(plot.background = element_rect(fill = "white", color = "white"))
      
      # Save output plot
      output_file <- here::here("data",
                          "figures",
                          paste0("02_2_combined_plot_", name, ".png"))
      ggsave(
        output_file,
        combined_plot,
        width = 20,
        height = 13,
        dpi = 300,
        bg = "white"
      )

      # Cleanup
      rm(merged_raster, windowed_data, correlation_df)
      gc()

      return(output_file)
    },
    .options = furrr::furrr_options(seed = TRUE)
  )

})

# Restore single-threaded processing
plan(sequential)
gc()
```

## Results and Interpretation

### (a) Water‐stressed environment of Arizona (Desert riparian)

```{r visuallization1}
img_paths <- results
knitr::include_graphics(img_paths[1])
```

### (b) Summer‐dry Mediterranean climate in central California


```{r visuallization2}
knitr::include_graphics(img_paths[2])
```

### (c) Forest‐savanna transition in Luama Katanga Reserve of eastern Congo


```{r visuallization3}
knitr::include_graphics(img_paths[3])
```

### (d) Seasonally waterlogged (oxygen‐stressed) Pantanal where larger vegetation occupies better drained mounds


```{r visuallization4}
knitr::include_graphics(img_paths[4])
```

### (e) White sand community in lowland Venezuela with both excessive drainage of the sand and waterlogging (Terborgh, 1992)

```{r visuallization5}
knitr::include_graphics(img_paths[5])
```

## Conclusion

This analysis involves cropping, merging, performing correlation analysis, and visualizing TWI and Tree Height data for various regions. Each step is carefully processed, and results are visualized in combined plots for each region.


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/02_2_splitting_window_correlation_validation.Rmd"))
```
