---
title: "Split Window Correlation -- Global Analysis. "
author: "Ting Tan"
date: "2025-05-05"
output:
  html_document:
    fig_align: 'center'
bibliography: literatures.bib  
latex_engine: xelatex
csl: apa.csl
---

```{r SetUp, include=FALSE, eval=TRUE}

knitr::opts_chunk$set(
  comment = "#>", echo = FALSE, fig.width = 10, message = FALSE, warning = FALSE
)

library(terra)
library(future)
library(tidyverse)
library(furrr)
library(scico)
library(viridis)
library(fs)
library(sf)
library(progressr)
library(tidyterra)
handlers(handler_txtprogressbar(style = 3))  

source(here::here("R/split_window_correlation.R"))
```

```{r Configuration}
file_vegh_450m_mosaic <- file.path("/data_2/scratch/ting/data/vegh_450m/vegh_450m_2020_mosaic.nc")
file_ga2 <- file.path("/data/archive/gti_marthews_2015/data/ga2.nc")  
cor_mosaic_file <- file.path("/data_2/scratch/ting/data/corr_map/corr_twi_vegh_5km_mosaic.nc")  
temp_dir <- file.path("/data_2/scratch/ting/data_temp")
cor_tiles_dir <- file.path("/data_2/scratch/ting/data/corr_map/30_30_deg")
ecoregions_path <- file.path("/data_2/scratch/ting/data_raw/ecoregion2017/Ecoregions2017/Ecoregions2017.shp")
```

---

## Objective

To explore the global pattern of vegetation structure in relation to hillslope hydrology, this study investigates how **vegetation height (VEGH)** varies with the **topographic wetness index (TWI)** using a window-based spatial correlation approach.

---

## Implementation

### 1. Data Preprocessing

* The global domain is divided into **30° × 30° tiles** to allow for efficient memory usage and parallel computation.
* Within each tile, the global **TWI** and **VEGH** raster mosaics are **cropped** and **merged** into a single NetCDF file for processing.
* All tile metadata—including tile name, geographic extent (`xmin`, `xmax`, `ymin`, `ymax`), and NetCDF file path—are saved as a tibble and written to a CSV file (`here::here("data/preprocessed_tile_info.csv")`).
* This structured metadata serves as the parameter input for parallel processing, thereby avoiding direct in-memory loading of large raster datasets, which can easily cause memory overflow.

### 2. Parallel Correlation Analysis

* Each tile is processed independently in parallel to increase computational efficiency.
* Tiles are binned into **12 × 12 pixel windows**, each covering approximately **0.05°** in spatial resolution.
* Within each window, a **Pearson correlation coefficient (r)** is calculated between TWI and VEGH values.
* The resulting correlation values are saved as raster tiles to:

  * `sftp://130.92.119.132/data_2/scratch/ting/data/corr_map/30_30_deg`
* Additionally, a **diagnostic figure** is generated for each tile, summarizing:

  * TWI and VEGH distributions,
  * spatial correlation results,
  * land cover overlays,
  * and scatterplots of TWI–VEGH relationships.
    Example figure path:
    `here::here("data/figures/02_3_n60_w90_vegh_and_twi_correlation_analysis.png")`

### 3. Mosaicking

* All individual tile-level correlation results are seamlessly **merged into a global mosaic**:

  * Output path: `sftp://130.92.119.132/data_2/scratch/ting/data/corr_map/corr_twi_vegh_5km_mosaic.nc`
* This mosaic serves as the basis for visualization and further analysis at the global scale.

---

## Visualization and Interpretation

* Global maps of **TWI**, **VEGH**, **Pearson correlation coefficients**, and **associated p-values** are visualized using `ggplot2` and `tidyterra`.
* Color scales are carefully selected (e.g., `scico`, `batlow`) to emphasize meaningful spatial variation.
* For interpretability:

  * Correlation maps use diverging color palettes centered at zero.
  * P-value maps use stepped thresholds to highlight significance levels (e.g., *p* < 0.05).
* In addition, combined diagnostic figures for each tile (from Step 2) are reviewed to identify local patterns and biome-specific interactions between vegetation structure and wetness conditions.

---

```{r Preprocess, eval=FALSE}
# once finished, no need for reprocessing (30 mins)
prep_tiles_info <- generate_prep_tiles_info(
  file_ga2, file_vegh_450m_mosaic, temp_dir)

str(prep_tiles_info, max.level = 2)
```


```{r Implementation, eval=FALSE}
# this process need more than 3h
prep_tiles_info <- readr::read_csv(here::here("data/preprocessed_tiles_info.csv"))

gc()
plan(multisession, workers = 6)

# Create a list to collect failed tiles
error_log <- list()

with_progress({
  pb <- progressor(along = prep_tiles_info)

  results <- future_pmap(
    prep_tiles_info,
    function(name, xmin, xmax, ymin, ymax, file) {
      tryCatch({
        
        pb(sprintf("Processing tile: %s", name))
        
        # Load pre-processed raster
        merged_raster <- terra::rast(file)
        
        # Generate windows and correlation dataframes
        windowed_data <- create_spatial_windows(merged_raster, 12)
        correlation_df <- calculate_window_correlations(windowed_data)
        
        # Convert to raster and Save the outcome
        correlation_raster <- terra::rast(
          correlation_df[, c("lon_cen", "lat_cen", "correlation", "cor_pval")], 
          type = "xyz", crs = "EPSG:4326")
        names(correlation_raster) <- c("correlation", "cor_pval")
        nc_path <- file.path(cor_tiles_dir, paste0(name, "_corr_twi_vegh_5km.nc"))
        terra::writeCDF(correlation_raster, cor_tiles_dir, overwrite = TRUE)
        
        # # Load processed raster
        # merged_raster <- terra::rast(file)
        # suppressWarnings({
        #   windowed_data <- as.data.frame(merged_raster, xy = TRUE, na.rm = TRUE)
        # })
        # colnames(windowed_data) <- c("lon", "lat", "twi", "vegh")
        # 
        # nc_path <- file.path(cor_tiles_dir, paste0("corr_twi_vegh_5km_", name, ".nc"))
        # corr_raster <- terra::rast(nc_path)
        # suppressWarnings({
        #   correlation_df <- as.data.frame(corr_raster, xy = TRUE, na.rm = TRUE)
        # })
        # colnames(correlation_df) <- c("lon_cen", "lat_cen", "correlation", "pval")
        
        # Generate plots
        plot1 <- plot_twi(windowed_data)
        plot2 <- plot_vegh(windowed_data)
        plot3 <- plot_corr(correlation_df)
        # plot4 <- plot_correlation_vs_pixel_count(correlation_df)
        plot5 <- plot_overview(windowed_data)
        
        sapply(list(plot1, plot2, plot3, plot5), class)
        
        source(here::here("R/split_window_correlation.R"))
        png_path <- save_combined_plot(
          plots = list(plot1, plot2, plot3, plot5),
          region_name = name,
          title_text = "VEGH and TWI Correlation Analysis",
          ncol = 2,
          width = 18,
          height = 16,
          file_index = "02_3"
        )
        
        # Clean memory
        rm(merged_raster,
           windowed_data,
           correlation_df,
           correlation_raster)
        gc()
        
        list(png_path)
      }, error = function(e) {
        msg <- sprintf("Tile %s failed: %s", name, e$message)
        message("❌ ", msg)
        error_log[[length(error_log) + 1]] <<- list(
          name = name,
          xmin = xmin,
          xmax = xmax,
          ymin = ymin,
          ymax = ymax,
          file = file,
          error = e$message
        )
        return(NA)
      })
    }
  )
  # Update progress bar in the main thread after all tasks are completed
  future_walk(results, ~pb()) 
})

# Restore single-threaded processing
plan(sequential)
gc()

# Summary
cat("✅ Completed tiles:", sum(!is.na(results)), "\n")
cat("❌ Failed tiles:", sum(is.na(results)), "\n")

# Save failed tiles log to CSV
if (length(error_log) > 0) {
  error_df <- bind_rows(error_log)
  log_file <- file.path(output_dir, "failed_tiles_log.csv")
  write_csv(error_df, log_file)
  message("📄 Error log saved to: ", log_file)
}

```


```{r Combination, eval=FALSE}
# combination all of the tiles with resolution of 5km (0.05°)
mosaic_and_save_rasters(cor_tiles_dir, cor_mosaic_file, verbose = TRUE)
```

```{r biomes, fig.width=12, fig.height=6}

suppressMessages(ecoregions <- st_read(ecoregions_path))

ggplot(data = ecoregions) +
  geom_sf(aes(fill = BIOME_NAME), color = NA) +
  scale_fill_manual(values = setNames(ecoregions$COLOR_BIO, ecoregions$BIOME_NAME)) +
  labs(title = "Biomes", fill = "Biome") +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-180, 180, by = 30)  # 每 30 度标注经度
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-60, 85),
    breaks = seq(-60, 90, by = 30)   # 每 30 度标注纬度
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 8)
  )

rm(ecoregions)
gc()
```
Figure 1. Biome distribution based on WWF ecoregions dataset, data from \citeauthor{dinersteinEcoregionBasedApproachProtecting2017} (\citeyear{dinersteinEcoregionBasedApproachProtecting2017}).

```{r twi, fig.width=12, fig.height=6}

twi_raster <- rast(file_ga2)

# plot TWI
ggplot() +
  tidyterra::geom_spatraster(data = twi_raster) +
  scale_fill_scico(palette = "oslo", direction = -1, na.value = "white") +
  labs(
    title = "Topographic Wetness Index (TWI /CTI)",
    fill = "TWI" ) +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-180, 180, by = 30)  # 每 30 度标注经度
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-60, 85),
    breaks = seq(-60, 90, by = 30)   # 每 30 度标注纬度
  ) +
  theme(legend.position = "right",
        legend.text = element_text(size = 6),
        legend.title = element_blank())

rm(twi_raster)
gc()
```
Figure 2. TWI indicates water accumulation tendency based on topography, data from \citeauthor{marthewsHighresolutionGlobalTopographic2015} (\citeyear{marthewsHighresolutionGlobalTopographic2015}).

```{r vegh, fig.width=12, fig.height=6}

vegh_raster <- rast(file_vegh_450m_mosaic)

# plot vegetation height
ggplot() +
  tidyterra::geom_spatraster(data = vegh_raster) +
  scale_fill_scico(palette = "batlow", direction = -1, na.value = "white") +
  labs(
    title = "Vegetation Height (VEGH)",
    fill = "VEGH"
  ) +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-180, 180, by = 30)  # 每 30 度标注经度
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-60, 85),
    breaks = seq(-60, 90, by = 30)   # 每 30 度标注纬度
  ) +
  theme(legend.position = "right",
        legend.text = element_text(size = 6),
        legend.title = element_blank())

rm(vegh_raster)
gc()
```
Figure 3. Vegetation Height (VEGH) data from \citeauthor{langHighresolutionCanopyHeight2023} (\citeyear{langHighresolutionCanopyHeight2023}).


```{r cor, fig.width=12, fig.height=6}

correlation_raster <- rast(cor_mosaic_file)

# plot pearson correlation
ggplot() +
  tidyterra::geom_spatraster(data = correlation_raster[[1]]) +
  scale_fill_scico(palette = "bam", direction = 1, na.value = "white") +
  labs(
    title = "VEGH-TWI Pearson Correlation",
    fill = "r") +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-180, 180, by = 30)  # 每 30 度标注经度
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-60, 85),
    breaks = seq(-60, 90, by = 30)   # 每 30 度标注纬度
  ) +
  theme(legend.position = "right",
        legend.text = element_text(size = 6),
        legend.title = element_blank())

```
Figure 4. VEGH–TWI Pearson correlation map. Input data from \citeauthor{langHighresolutionCanopyHeight2023} (\citeyear{langHighresolutionCanopyHeight2023}) and \citeauthor{marthewsHighresolutionGlobalTopographic2015} (\citeyear{marthewsHighresolutionGlobalTopographic2015}).

  
```{r p_val, fig.width=12, fig.height=6}
# plot p-value of the correlation analysis
ggplot() +
  tidyterra::geom_spatraster(data = correlation_raster[[2]]) +
  scale_fill_stepsn(
    colours = c("#2166AC", "#67A9CF", "#D1E5F0", "#FDDBC7", "#EF8A62", "#B2182B"),
    values = scales::rescale(c(0, 0.01, 0.05, 0.1, 0.5, 1)),  # 分段值（rescaled to 0–1）
    breaks = c(0.01, 0.05, 0.1, 0.5, 1),
    limits = c(0, 1),
    oob = scales::squish,
    name = "p",
    na.value = "white"
  ) +
  labs(
    title = "P-value") +
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(-180, 180, by = 30)  # 每 30 度标注经度
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(-60, 85),
    breaks = seq(-60, 90, by = 30)   # 每 30 度标注纬度
  ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 8)
  )
rm(correlation_raster)
gc()
```
Figure 5. The P-value of the Pearson correlation analysis between TWI and VEGH. Input data from \citeauthor{langHighresolutionCanopyHeight2023} (\citeyear{langHighresolutionCanopyHeight2023}) and \citeauthor{marthewsHighresolutionGlobalTopographic2015} (\citeyear{marthewsHighresolutionGlobalTopographic2015}).


```{r Visulization2}
png_files <- list.files(
  path = here::here("data/figures"),                 
  pattern = "_vegh_and_twi_correlation_analysis\\.png$", 
  full.names = TRUE                     
)

knitr::include_graphics(png_files)
```

```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/02_3_splitting_window_correlation_global.Rmd"))
```
