---
title: "Additional Analysis of Natural Lands with Minimal Land Use"
author: "Ting Tan"
date: "2025-08-19"
output:
  html_document:
    fig_align: 'center'
bibliography: literatures.bib  
latex_engine: xelatex
csl: apa.csl
---

```{r SetUp, include=FALSE, eval=TRUE}

library(terra)
library(dplyr)
library(ggplot2)
library(viridis)
library(stringr)

# Load configuration and functions
source(here::here("config.R"))
```

---

## Introduction

Topography shapes vegetation patterns through two major pathways:

1. **Water availability** (groundwater table depth, soil moisture), which creates **dry ridges** and **moist valley bottoms**.
2. **Energy conditions** (solar radiation, evapotranspiration), which generate **sunny slopes** and **shady slopes**.

From the global correlation map between vegetation height (H) and the Topographic Wetness Index (TWI), it is evident that topography strongly regulates vegetation growth by altering local water regimes. *At the global scale, negative correlations between H and TWI are more widespread and stronger in magnitude.** This suggests that in many climatic regions, topographic water accumulation tends to produce **overly wet or light-limited environments**, thereby constraining vegetation height. By contrast, in arid and semi-arid regions, H and TWI show significant positive correlations, reflecting the **facilitative role of water convergence** under water-limited conditions.

The spatial pattern of these correlations aligns closely with biome distributions, indicating that the **direction and strength of topographic control are climate-dependent**:

* **Humid regions:** predominantly negative effects, where excess soil water and poor aeration suppress growth.
* **Arid and semi-arid regions:** predominantly positive effects, where local water convergence enhances growth.

This finding highlights that vegetation patterns emerge from the **interaction among climate, topography, and vegetation**.

## Task

* **Biome-wise statistics:** Quantify the distribution of H–TWI correlations within each biome (boxplots or probability density curves) for intuitive comparisons.
* **Visualization:** Produce a global correlation map alongside biome-specific histograms to emphasize differences in both **direction** and **magnitude**.
* **Interpretation framework:** Situate the results within the **water–energy compensation hypothesis**, which posits that water and energy act as complementary limiting factors—when water is abundant, energy becomes limiting; when energy is sufficient, water becomes limiting. This framing adds ecological significance to the observed patterns.

---

```{r}
cor_natura <- rast(cor_twi_vegh_mask_fused0.05_file)
mat_5km_r <- rast(mat_5km_file)
ecoregion_r <- rast(ecoregion_5km_path)

stacked <- c(cor_natura, mat_5km_r, ecoregion_r)

# Convert to data frame 
df <- as.data.frame(stacked, xy = FALSE, na.rm = TRUE)
colnames(df) <- c("cor", "mat", "ecoregion")

```

### the scatterplot of r(H~TWI) VS. MAT 

this is almost the same with  r(H~TWI) VS. AI.
maybe we could do  r(H~TWI) VS. P 
  see Adeney et al. (2016) White-sand Ecosystems in Amazonia (P9)

```{r}
ggplot(df, aes(x = mat, y = cor)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(option = "C", trans = "log1p") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1) +
  # coord_cartesian(xlim = c(0, 7), ylim = c(-1, 1)) +  # 保留所有数据，但限制显示范围
  labs(
    title = "r(H~TWI) VS. MAT",
    x = "Mean Annual Temperature",
    y = "r(H~TWI)",
    fill = "Count"
  ) +
  theme_bw() +
  theme(legend.position = "none")

```

### get the lable and color for each BIOME

```{r}
ecoregion <- vect(ecoregion_path)
biomes <- ecoregion |>
  as.data.frame() |>
  dplyr::select(BIOME_NUM, BIOME_NAME, COLOR_BIO) |>
  dplyr::distinct() |>
  dplyr::arrange(BIOME_NUM)

biome_colors <- setNames(biomes$COLOR_BIO, biomes$BIOME_NUM)
biome_labels <- setNames(biomes$BIOME_NAME, biomes$BIOME_NUM)

```



### plot violin and boxplot for each ecoregion

```{r,fig.width=12,fig.height=8}
df_biome <-  df |>
  group_by(ecoregion)

ggplot(df_biome, aes(x = factor(ecoregion, levels = names(biome_labels), labels = biome_labels), 
                     y = cor, 
                     fill = factor(ecoregion))) +
  geom_violin(alpha = 0.6, trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(
    values = biome_colors,
    labels = biome_labels,
    name   = "Biome"
  ) +
  labs(
    x = "Ecoregion",
    y = "r(H~TWI)",
    title = "Distribution of r(H~TWI) by Ecoregion"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 0, 
      hjust = 0.5, 
      size = 8
    ),
    legend.position = "none"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) # 自动换行 X 轴标签，每行最多 15 个字符。
```


### plot density for each ecoregion

```{r,fig.width=12,fig.height=8}
biome_colors_named <- setNames(biomes$COLOR_BIO, biomes$BIOME_NAME)
df$ecoregion_name <- biomes$BIOME_NAME[match(df$ecoregion, biomes$BIOME_NUM)]
df$ecoregion_name <- factor(df$ecoregion_name, levels = names(biome_colors_named))

ggplot(df, aes(x = cor, fill = ecoregion_name, color = ecoregion_name)) +
  geom_density(alpha = 0.2) +
  scale_fill_manual(values = biome_colors_named) +
  scale_color_manual(values = biome_colors_named) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "r(H~TWI)", y = "Density", title = "Density Distribution of r(H~TWI) by Biome") +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(
    color = guide_legend(nrow = 5, byrow = TRUE),  # 图例分 3 行
    fill  = guide_legend(nrow = 5, byrow = TRUE)
  )

```


```{r,fig.width=12,fig.height=8}
ggplot(df, aes(x = cor, fill = ecoregion_name, color = ecoregion_name)) +
  geom_density(alpha = 0.3, size = 0.8) +
  scale_fill_manual(values = biome_colors_named) +
  scale_color_manual(values = biome_colors_named) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ ecoregion_name, ncol = 4, scales = "free_y",
             labeller = label_wrap_gen(width = 15)) +  # 自动换行
  labs(
    x = "r(H~TWI)",
    y = "Density",
    title = "Density Distribution of r(H~TWI) by Biome"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8),
    strip.text = element_text(size = 9),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

```


---

## Interpretation and Implications

Overall, the results are broadly consistent with our hypothesis, indicating that **systematic topographic control of vegetation height does exist at the global scale**. However, to strengthen the signal, it will be important to **reduce noise**, particularly by addressing windows that show **non-monotonic correlations** or are based on **limited observations**.

The biome-level patterns can be summarized as follows:

1. **Strong Negative Relationships**

   * Tropical & Subtropical Moist Broadleaf Forests
   * Temperate Broadleaf & Mixed Forests
   * Mangroves
   * Flooded Grasslands & Savannas

   → In these humid ecosystems, topographic water convergence often leads to **overly wet and light-limited conditions**, suppressing vegetation growth.

2. **Weak Negative Relationships**

   * Tropical & Subtropical Coniferous Forests
   * Boreal Forests/Taiga

   → Waterlogging still constrains vegetation, but the effect is less pronounced, likely due to cold climates or adaptations of coniferous vegetation.

3. **Strong Positive Relationships**

   * Montane Grasslands & Shrublands
   * Tropical & Subtropical Dry Broadleaf Forests
   * Temperate Grasslands, Savannas & Shrublands
   * Tropical & Subtropical Grasslands, Savannas & Shrublands

   → In arid and semi-arid systems, **local water accumulation enhances growth**, producing a clear positive correlation between topography and vegetation height.

4. **Weak Positive Relationships**

   * Temperate Conifer Forests
   * Tundra
   * Mediterranean Forests, Woodlands & Scrub
   * Deserts & Xeric Shrublands

   → Positive tendencies are present but weaker, suggesting that other limiting factors (cold temperatures, shallow soils, or extreme drought) reduce the relative importance of topography.

---

## Improvement 

1. **Noise reduction:**

   * Exclude windows with very few observations (set a minimum N threshold).
   * Apply spatial filtering (median smoothing across neighboring windows).
   * Consider masking out extreme peak values (robust z-score filter).

2. **Window definition:**

   * Test sensitivity to different moving window sizes (e.g., 5 km vs 10 km).
   * Check whether strong peaks are artifacts of small sample size.

3.  **Nonlinear and unimodal relationships:**

   * Identify windows where H–TWI relationships deviate from monotonic trends.
   * Fit nonlinear models (e.g., quadratic regression, GAMs, or segmented regression) to capture unimodal responses.
   * Use metrics such as variance explained (R²) or model comparison to assess whether nonlinear fits better describe the data than simple correlations.
   * Treat unimodal windows separately in interpretation, as they may indicate **thresholds or optimal ranges** of water availability rather than purely positive or negative effects.



```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/03_re3_r_in_natural_land.Rmd"))
```
