---
title: "Pearson's r (H~TWI) Calculation in Sample Regions"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

## Task

The Pearson correlation analysis between vegetation height and topographic wetness index for all automaticlly selected region has been completed in `~/veg_topo/analysis/04_3_regional_twi_vegh_cor.R`.

This markdown file focuses on visualizing the results, generating plots for each tile for easier inspection and interpretation.

### hyper_arid_flat_relief

```{r}

# list all figures of global correlation (H~TWI) for all 49 tiles
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_hyper_arid_flat_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### hyper_arid_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_hyper_arid_rugged_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### semiarid_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_semiarid_flat_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### semiarid_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_semiarid_rugged_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### dry_subhumid_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_dry_subhumid_flat_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### dry_subhumid_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_dry_subhumid_rugged_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### moist_subhumid_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_moist_subhumid_flat_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```


### moist_subhumid_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_moist_subhumid_rugged_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### perhumid_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_perhumid_flat_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### perhumid_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^04_regA_perhumid_rugged_relief.*_win_500m_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/04_re3.Rmd"))
```

