---
title: "Regions selection for topographic effect analysis"
author: "Ting Tan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
    fig_align: 'center'
bibliography: literatures.bib  
latex_engine: xelatex
csl: apa.csl
---


```{r SetUp, include=FALSE, eval=TRUE}

knitr::opts_chunk$set(
  comment = "#>", fig.width = 10, message = FALSE, warning = FALSE
)

library(terra)
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(DataExplorer)
library(RColorBrewer)
library(rnaturalearth)
library(sf)

# Load configuration and functions
source(here::here("config.R"))

```


---

## Objective

This study aims to investigate the **topographic effects on vegetation patterns**, which may exhibit different signals across spatial scales. At the **regional scale**, we aggregate the raw 30 m resolution data into 20 × 20 pixel windows and calculate the window-based Pearson correlation between the **Topographic Wetness Index (TWI)** and **vegetation height**. However, due to the high computational demands, it is not feasible to apply this method at the global scale. Therefore, we selectively focus on a few representative regions for detailed analysis. To minimize the influence of human activities on natural vegetation patterns, we prioritize regions that are as undisturbed as possible.

Under the framework of Fan et al. \[@fanHillslopeHydrologyGlobal2019, year-only], **topographic controls on vegetation** primarily operate through two mechanisms: **lateral subsurface flow** and **solar insolation** (via slope aspect). The interaction between hydroclimate and topographic relief determines the extent and direction of lateral flow impacts on vegetation growth. Similarly, variations in latitude and relief influence insolation patterns and, in turn, vegetation response.

Accordingly, we select 12 representative regions for analysis, grouped by the dominant topographic control.

### Lateral Subsurface Flow-Dominated Patterns

(Selected based on hydroclimate and relief conditions):

1. Very dry hydroclimate with flat relief
2. Moderate hydroclimate with flat relief
3. Very wet hydroclimate with flat relief
4. Very dry hydroclimate with rugged relief
5. Moderate hydroclimate with rugged relief
6. Very wet hydroclimate with rugged relief

### Insolation-Dominated Patterns

(Selected based on latitude and relief conditions):

1. Very low latitude with flat relief
2. Moderate latitude with flat relief
3. Very high latitude with flat relief
4. Very low latitude with rugged relief
5. Moderate latitude with rugged relief
6. Very high latitude with rugged relief

---

## Implementation

### Avaliable Data

```{r DataPre}
# Load resampled raster datasets (AI, TWI, fused)

ai_55km_r <- terra::rast(ai_55km_file) * 0.0001 # Multiply all values by 0.0001 to get original value
fused_55km_r <- terra::rast(fused_55km_file)
dem_sd_55km_r <- terra::rast(dem_sd_55km_path)
fpa_55km_r <- terra::rast(fpa_55km_path)
fvegh_55km_r <- terra::rast(fvegh_55km_path)

# Stack rasters into a single SpatRaster
stacked <- c(ai_55km_r,
             fused_55km_r,
             dem_sd_55km_r,
             fpa_55km_r,
             fvegh_55km_r)

# Use land polygons to remove ocean area with NA

# Load land polygons
land <- rnaturalearth::ne_countries(scale = 110, returnclass = "sf")
land_vect <- vect(land)

# Crop and mask raster to land
stacked_masked <- mask(crop(stacked, land_vect), land_vect)

plot(stacked_masked)

# Convert to data frame
df <- as.data.frame(stacked_masked, xy = TRUE, na.rm = TRUE)
colnames(df) <- c("lon", "lat", "ai", "fused", "relief", "fpa", "fvegh")

# Clean up
rm(ai_55km_r, fused_55km_r, dem_sd_55km_r, fpa_55km_r, fvegh_55km_r)
gc()

```


```{r DataOverview}
# Ploting
p_hg <- DataExplorer::plot_histogram(df)

# combine the plots
cp_hg <- patchwork::wrap_plots(p_hg)

# Save histogram plot
ggsave(
  filename = here::here("data/figures/04_variables_hg.png"),
  plot = cp_hg,
  width = 8,
  height = 6,
  dpi = 300,
  units = "in"
)
```


### Create bins

```{r Bin, echo = TRUE}
df_binned <- df |>

  # create bins
  mutate(
    ai_bin = cut(ai, breaks = quantile(ai, probs = seq(0, 1, 1/5)), include.lowest = TRUE),
    relief_bin = cut(relief, breaks = quantile(relief, probs = seq(0, 1, 1/3)), include.lowest = TRUE),
    abs_lat = abs(lat),
    abs_lat_bin = cut(abs_lat, breaks = quantile(abs_lat, probs = seq(0, 1, 1/5)), include.lowest = TRUE)
  ) |>

  # Fig. 6 Down Valley drainage
  mutate(strata_A = interaction(ai_bin, relief_bin, drop = TRUE)) |>

  # Fig. 6 Aspect difference
  mutate(strata_B = interaction(abs_lat_bin, relief_bin, drop = TRUE)) |>

  # Assign readable labels to each bin
  mutate(
    ai_label = case_when(
      ai_bin == levels(ai_bin)[1] ~ "very_dry",
      ai_bin == levels(ai_bin)[2] ~ "dry",
      ai_bin == levels(ai_bin)[3] ~ "moderate",
      ai_bin == levels(ai_bin)[4] ~ "humid",
      ai_bin == levels(ai_bin)[5] ~ "very_humid"
    ),
    relief_label = case_when(
      relief_bin == levels(relief_bin)[1] ~ "flat_relief",
      relief_bin == levels(relief_bin)[2] ~ "moderate_relief",
      relief_bin == levels(relief_bin)[3] ~ "rugged_relief"
    ),
    lat_label = case_when(
      abs_lat_bin == levels(abs_lat_bin)[1] ~ "low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[2] ~ "relatively_low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[3] ~ "mid_latitude",
      abs_lat_bin == levels(abs_lat_bin)[4] ~ "relatively_high_latitude",
      abs_lat_bin == levels(abs_lat_bin)[5] ~ "high_latitude"
    )
  ) |>

  # Combine text labels into new readable strata labels
  mutate(
    strata_A_label = paste0(ai_label, "_", relief_label),
    strata_B_label = paste0(lat_label, "_", relief_label)
  ) |>

  select(-ai_label, -relief_label, -lat_label)

# check data
table(df_binned$ai_bin)
table(df_binned$abs_lat_bin)
table(df_binned$relief_bin)

table(df_binned$strata_A_label)
table(df_binned$strata_B_label)

```

### Region filter

#### fused < 0.05, fpa > 0.8, fvegh > 0.8

The criteria fused < 0.05 and fpa > 0.8 tend to select regions dominated by water bodies, bare ground, or other sparsely vegetated surfaces. To ensure that the selected regions are suitable for analyzing natural vegetation patterns, we introduce an additional condition: the region must have a vegetation data fraction greater than 0.8 (fvegh > 0.8).

However, this strict combination of filters may exclude many areas of interest—particularly flat, mid-latitude regions—since such landscapes are often heavily developed or used for agriculture and therefore lack large protected areas.



```{r Filter, echo = TRUE}

df_filt_A <- df_binned |>

  filter(fused < 0.05, fpa > 0.9, fvegh > 0.8) |>
  
  filter(
    ai_bin %in% levels(ai_bin)[c(1, 3, 5)], # bin 1 are too dry that has little fraction of vegetation
    relief_bin %in% levels(relief_bin)[c(1, 3)],
  )

table(df_filt_A$strata_A_label)

df_filt_B <- df_binned |>

  filter(fused < 0.05, fpa > 0.9, fvegh > 0.8) |>
  
  filter(
    relief_bin %in% levels(relief_bin)[c(1, 3)],
    abs_lat_bin %in% levels(abs_lat_bin)[c(1, 3, 5)]
  )

table(df_filt_B$strata_B_label)
```


#### Very Dry Areas with Naturally Low Vegetation Cover

Very dry regions often have inherently low fractions of green vegetation, which may cause them to be excluded by strict vegetation thresholds. To address this limitation, we propose including certain arid regions that meet a relaxed criterion of fvegh > 0.5. This approach helps maintain geographic and ecological representativeness in our analysis.

```{r dry}
df_dry <- df_binned |>
  filter(ai_bin %in% levels(ai_bin)[1]) |>
  select(fpa, fvegh, fused, ai_bin) 

summary(df_dry)

df_filt_A_dry <- df_binned |>

  filter(fused < 0.05, fpa > 0.8, fvegh > 0.5) |> # fvegh > 0.5
  
  filter(
    ai_bin %in% levels(ai_bin)[c(1)], 
    relief_bin %in% levels(relief_bin)[c(1)],
  )

table(df_filt_A_dry$strata_A_label)
```

#### Lack of Large Protected Areas in Flat Mid-Latitude Regions

Large protected areas are generally absent in flat, mid-latitude regions, as these areas are often prime locations for development and agriculture. To address this limitation, we propose including certain flat mid-latitude regions that meet a relaxed criterion of `fpa > 0.5`. This adjustment ensures better geographic and ecological representativeness in our analysis.

```{r mid_flat}
# mid_lat_flat_relief (32.5° - 49°)
# have vey low fpa !!!!!!!!!
df_mid_flat <- df_binned |>
  filter(strata_B_label == "mid_latitude_flat_relief") |>
  select(fpa, fvegh, fused, abs_lat_bin, relief_bin) 

  summary(df_mid_flat)
  

df_filt_B_mid_flat <- df_binned |>

  filter(fused < 0.05, fpa > 0.5, fvegh > 0.8) |> # fpa > 0.5
  
  filter(
    relief_bin %in% levels(relief_bin)[c(1)],
    abs_lat_bin %in% levels(abs_lat_bin)[c(3)]
  )

table(df_filt_B_mid_flat$strata_B_label)
```



```{r Filterbind, echo = TRUE}

df_filt_A_all <- df_filt_A |>
  bind_rows(df_filt_A_dry)

table(df_filt_A_all $strata_A_label)

df_filt_B_all <- df_filt_B |>
  bind_rows(df_filt_B_mid_flat)

table(df_filt_B_all $strata_B_label)
```

---

## Visuallization 

```{r}

# load coast outline, vector data
coast <- rnaturalearth::ne_coastline(scale = 110, returnclass = "sf")

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_A_all, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_B_all, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()
```


---

## Sampling

```{r Sampling}
# Pick one region (coarse gridcell) representing all combination of bins.
# Should yield 6 samples

set.seed(999)

# for Fig. 6 Down Valley drainage
df_samples_A <- df_filt_A_all |>
  group_by(strata_A) |>
  sample_n(size = 1, replace = FALSE) |>
  mutate(
    xmin = lon - 0.25,
    xmax = lon + 0.25,
    ymin = lat - 0.25,
    ymax = lat + 0.25
  ) |>
  select(-starts_with("strata_B")) |>
  ungroup()

# for Fig. 6 Aspect difference
df_samples_B <- df_filt_B_all |>
  group_by(strata_B) |>
  sample_n(size = 1, replace = FALSE) |>
  mutate(
    xmin = lon - 0.25,
    xmax = lon + 0.25,
    ymin = lat - 0.25,
    ymax = lat + 0.25
  ) |>
  select(-starts_with("strata_A")) |>
  ungroup()


# Save the two data frames as RDS files
saveRDS(df_samples_A, here::here("data/df_samples_A.rds"))
saveRDS(df_samples_B, here::here("data/df_samples_B.rds"))
```

```{r gpkg}
# ---- regA
# load region infos
regA_info <- readRDS(here::here("data/df_samples_A.rds")) |>
  select(ends_with("label"), ends_with("min"), ends_with("max"))

regA_gpkg <- regA_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regA_gpkg, here::here("data/regionA_boxes.gpkg"), delete_layer = TRUE)


# ---- regB
# load region infos
regB_info <- readRDS(here::here("data/df_samples_B.rds")) |>
  select(ends_with("label"), ends_with("min"), ends_with("max"))

regB_gpkg <- regB_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regB_gpkg, here::here("data/regionB_boxes.gpkg"), delete_layer = TRUE)
```
