---
title: "Regions selection for topographic effect analysis"
author: "Ting Tan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
    fig_align: 'center'
bibliography: literatures.bib  
latex_engine: xelatex
csl: apa.csl
---

```{r SetUp, include=FALSE, eval=TRUE}

knitr::opts_chunk$set(
  comment = "#>", fig.width = 10, message = FALSE, warning = FALSE
)

library(terra)
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(DataExplorer)
library(RColorBrewer)
library(rnaturalearth)
library(sf)

# Load configuration and functions
source(here::here("config.R"))
```

---

## **Introduction**

Topography redistributes subsurface water, producing wetter valleys that often support higher vegetation productivity and drier hilltops where growth is constrained. Conversely, excessive subsurface water (e.g., shallow water tables) can suppress vegetation via waterlogging and oxygen stress. Topographic position—commonly summarized by the Topographic Wetness Index (TWI)—is therefore a key control on vegetation patterns through its regulation of subsurface moisture. The strength and sign of this control, however, depend strongly on hydroclimatic context. Global analyses show that the correlation between TWI and vegetation height exhibits spatial structure that aligns with aridity gradients and plant functional types. Human activities (land use, irrigation, logging, etc.) can substantially alter or mask these natural relationships.

Topography also produces strong aspect-driven contrasts (sunny versus shady slopes), which modify the local energy and water balance and thus vegetation productivity. The magnitude and seasonal timing of aspect effects are modulated by latitude and relief: the same slope aspect produces very different insolation regimes in the tropics versus high latitudes, and in gentle versus rugged terrain.

To isolate the natural topography–vegetation relationships at fine spatial scales, and to validate the framework of Fan et al. (2019), we focus on regions with minimal human management where vegetation patterns most clearly reflect topographic hydrological and insolation controls. Specifically, we examine how climate–terrain combinations shape down-valley drainage and hydrological isolation (TWI-driven effects), and how latitude–terrain combinations modulate slope-aspect (insolation-driven) effects.

---

## **Objective**

Automatically identify regions where vegetation patterns are predominantly terrain-driven across different hydroclimatic and latitudinal backgrounds, in order to validate the Fan et al. (2019) framework. This requires masking or excluding areas where vegetation is primarily controlled by non-topographic drivers, including intensive human management, extreme temperatures, persistent snow or ice, very high wind exposure, coastal or tidal effects, and locations with poor or unreliable data.

## Pattern taxonomy

Two complementary control sets are used for site selection and testing:

### **A. Lateral Subsurface Flow Patterns**

(Defined by hydroclimate and relief conditions)

1. Very dry hydroclimate + flat relief
2. Very dry hydroclimate + rugged relief
3. Moderate hydroclimate + flat relief
4. Moderate hydroclimate + rugged relief
5. Very wet hydroclimate + flat relief
6. Very wet hydroclimate + rugged relief

### **B. Insolation Patterns**

(Defined by latitude and relief conditions)

1. Very low latitude + flat relief
2. Very low latitude + rugged relief
3. Mid latitude + flat relief
4. Mid latitude + rugged relief
5. Very high latitude + flat relief
6. Very high latitude + rugged relief

---

## Implementation

### Available Data

We use globally resampled raster datasets at 55 km resolution, including:

* **Aridity Index (AI)** – scaled to original values.
* **Fused human influence index** (fused) – measures combined human footprint.
* **Fraction of water/ice (FWI)** – to exclude non‐vegetated or cold regions.
* **Relief (standard deviation of elevation)** and **elevation range** from DEM.
* **Fraction of protected area (FPA)** – to select minimally disturbed regions.
* **Fractional vegetation height (fVegH)** from Lang et al. (2023).
* **Mean annual temperature (MAT)** – to avoid cold‐limited regions.

These datasets are stacked into a single `SpatRaster`, masked to land areas using `rnaturalearth` polygons, and converted into a tidy data frame with longitude, latitude, and variables of interest.

```{r DataPre}
# Load resampled raster datasets (AI, TWI, fused)

ai_55km_r <- terra::rast(ai_55km_file) * 0.0001 # Multiply all values by 0.0001 to get original value
fused_55km_r <- terra::rast(fused_55km_file)
fwi_55km_r <- terra::rast(fwi_55km_file)
dem_sd_55km_r <- terra::rast(dem_sd_55km_path)
dem_rg_55km_r <- terra::rast(dem_rg_55km_path)
fpa_55km_r <- terra::rast(fpa_55km_path)
fvegh_55km_r <- terra::rast(fvegh_55km_path)
mat_55km_r <- terra::rast(mat_55km_file)

# Stack rasters into a single SpatRaster
stacked <- c(ai_55km_r,
             fused_55km_r,
             fwi_55km_r,
             dem_sd_55km_r,
             dem_rg_55km_r,
             fpa_55km_r,
             fvegh_55km_r,
             mat_55km_r)

# Use land polygons to remove ocean area with NA

# Load land polygons
land <- rnaturalearth::ne_countries(scale = 110, returnclass = "sf")
land_vect <- vect(land)

# Crop and mask raster to land
stacked_masked <- mask(crop(stacked, land_vect), land_vect)

plot(stacked_masked)

# Convert to data frame
df <- as.data.frame(stacked_masked, xy = TRUE, na.rm = TRUE)
colnames(df) <- c("lon", "lat", "ai", "fused", "fwi", "relief", "dem_rg", "fpa", "fvegh", "mat")

# Clean up
rm(ai_55km_r, fused_55km_r, fwi_55km_r, dem_sd_55km_r, dem_rg_55km_r, fpa_55km_r, fvegh_55km_r, mat_55km_r)
gc()


# load coast outline, vector data for plotting
coast <- rnaturalearth::ne_coastline(scale = 110, returnclass = "sf")

```


```{r DataOverview}
# Ploting
p_hg <- DataExplorer::plot_histogram(df)

# combine the plots
cp_hg <- patchwork::wrap_plots(p_hg)

# Save histogram plot
ggsave(
  filename = here::here("data/figures/04_variables_hg.png"),
  plot = cp_hg,
  width = 8,
  height = 6,
  dpi = 300,
  units = "in"
)
```
---

### Binning Variables

To classify the global landscape into meaningful categories, we create bins for key drivers:

* **Aridity index (AI)**: split into five equal‐frequency bins (very dry → very humid).
* **Relief**: split into three equal‐frequency bins (flat → rugged).
* **Absolute latitude**: split into five equal‐frequency bins (low → high latitude).

From these, we define two sets of strata:

1. **Strata A** – combination of AI bin and relief bin (for lateral subsurface flow patterns).
2. **Strata B** – combination of latitude bin and relief bin (for insolation‐driven patterns).

Readable text labels (e.g., `"very_dry_flat_relief"`) are assigned for easier interpretation.

```{r Bin, echo = TRUE}
df_binned <- df |>

  # create bins
  mutate(
    ai_bin = cut(ai, breaks = quantile(ai, probs = seq(0, 1, 1/5)), include.lowest = TRUE),
    relief_bin = cut(relief, breaks = quantile(relief, probs = seq(0, 1, 1/3)), include.lowest = TRUE),
    abs_lat = abs(lat),
    abs_lat_bin = cut(abs_lat, breaks = quantile(abs_lat, probs = seq(0, 1, 1/5)), include.lowest = TRUE)
  ) |>

  # Fig. 6 Down Valley drainage
  mutate(strata_A = interaction(ai_bin, relief_bin, drop = TRUE)) |>

  # Fig. 6 Aspect difference
  mutate(strata_B = interaction(abs_lat_bin, relief_bin, drop = TRUE)) |>

  # Assign readable labels to each bin
  mutate(
    ai_label = case_when(
      ai_bin == levels(ai_bin)[1] ~ "very_dry",
      ai_bin == levels(ai_bin)[2] ~ "dry",
      ai_bin == levels(ai_bin)[3] ~ "moderate",
      ai_bin == levels(ai_bin)[4] ~ "humid",
      ai_bin == levels(ai_bin)[5] ~ "very_humid"
    ),
    relief_label = case_when(
      relief_bin == levels(relief_bin)[1] ~ "flat_relief",
      relief_bin == levels(relief_bin)[2] ~ "moderate_relief",
      relief_bin == levels(relief_bin)[3] ~ "rugged_relief"
    ),
    lat_label = case_when(
      abs_lat_bin == levels(abs_lat_bin)[1] ~ "low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[2] ~ "relatively_low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[3] ~ "mid_latitude",
      abs_lat_bin == levels(abs_lat_bin)[4] ~ "relatively_high_latitude",
      abs_lat_bin == levels(abs_lat_bin)[5] ~ "high_latitude"
    )
  ) |>

  # Combine text labels into new readable strata labels
  mutate(
    strata_A_label = paste0(ai_label, "_", relief_label),
    strata_B_label = paste0(lat_label, "_", relief_label)
  ) |>

  select(-ai_label, -relief_label, -lat_label)

# check data
table(df_binned$ai_bin)
table(df_binned$abs_lat_bin)
table(df_binned$relief_bin)

table(df_binned$strata_A_label)
table(df_binned$strata_B_label)
```

---

### Filtering Candidate Regions

#### For Lateral Subsurface Flow–Dominated Analysis (Strata A)

We retain only regions that meet all of the following:

* AI in very dry, moderate, or very humid categories.
* Relief in flat or rugged categories.
* Low human footprint (`fused < 0.05`) and minimal water/ice cover (`fwi < 0.1`).
* Sufficient vegetation height data (`fVegH > 0.1`).
* High protection status (`FPA > 0.9`).
* Warm climate (`MAT > 10 °C`).
* Moderate elevation range (< 500 m).

```{r FilterA, echo = TRUE}

df_filt_A <- df_binned |>
  
  filter(ai_bin %in% levels(ai_bin)[c(1, 3, 5)],
         relief_bin %in% levels(relief_bin)[c(1, 3)],) |>
  
  filter(fused < 0.05, fwi < 0.1, ) |>  # Exclude the region with high human activity and high invalid or cold areas (such as water, ice, or snow).
  
  filter(fvegh > 0.1) |> # avoid poor data from vegh data of Lang et.al (2023)
  
  filter(fpa > 0.5) |> # slect regions with 90% protected areas
  
  filter(mat > 10) |>       # avoid cold regions for hydrology-specific analysis
  
  filter(dem_rg < 500)   # consider only gridcells where the range of elevation is less than 500 m.

# print the numbers of possible region
table(df_filt_A$strata_A_label)
```

#### For Insolation‐Dominated Analysis (Strata B)

We retain only regions with:

* Low human footprint and high protection status.
* Relief in flat or rugged categories.
* Latitude in very low, mid, or very high categories.

```{r FilterB, echo = TRUE}

df_filt_B <- df_binned |>

  filter(fused < 0.05, fpa > 0.5, fwi < 0.1 ) |>
  
  filter(
    relief_bin %in% levels(relief_bin)[c(1, 3)],
    abs_lat_bin %in% levels(abs_lat_bin)[c(1, 3, 5)]
  )

# print the numbers of possible region
table(df_filt_B$strata_B_label)
```

---

### Sampling Representative Locations

From each set of filtered strata, we randomly select **one coarse grid cell** to represent each category combination. This yields:

* **6 samples** for lateral subsurface flow analysis (Strata A).
* **6 samples** for insolation‐driven analysis (Strata B).

For each sample, we define a bounding box (±0.25° in both latitude and longitude) and record the central coordinates.

```{r Sampling}
# Pick one region (coarse gridcell) representing all combination of bins.
# Should yield 6 samples

set.seed(2222)

# for Fig. 6 Down Valley drainage
df_samples_A <- df_filt_A |>
  group_by(strata_A) |>
  sample_n(size = 3, replace = FALSE) |>
  mutate(
    xmin = lon - 0.25,
    xmax = lon + 0.25,
    ymin = lat - 0.25,
    ymax = lat + 0.25,
    xmid= (xmin + xmax)/2,
    ymid = (ymin + ymax)/2,
    sample_id = row_number()
  ) |>
  select(-starts_with("strata_B")) |>
  ungroup()

# for Fig. 6 Aspect difference
df_samples_B <- df_filt_B |>
  group_by(strata_B) |>
  sample_n(size = 3, replace = FALSE) |>
  mutate(
    xmin = lon - 0.25,
    xmax = lon + 0.25,
    ymin = lat - 0.25,
    ymax = lat + 0.25,
    xmid= (xmin + xmax)/2,
    ymid = (ymin + ymax)/2,
    sample_id = row_number()
  ) |>
  select(-starts_with("strata_A")) |>
  ungroup()


# Save the two data frames as RDS files
saveRDS(df_samples_A, here::here("data/df_samples_A.rds"))
saveRDS(df_samples_B, here::here("data/df_samples_B.rds"))
```

---

### Export and Visualization

* Bounding boxes are saved as GeoPackage (`.gpkg`) files for GIS use.
* Locations are plotted on global maps, color‐coded by strata category.
* Additional maps show the global distribution of bins, filtered areas, and final selected regions.

```{r loadinfo}
# load region infos
regA_info <- readRDS(here::here("data/df_samples_A.rds")) |>
  select(ends_with("label"), ends_with("min"), ends_with("max"), ends_with("mid"))
regB_info <- readRDS(here::here("data/df_samples_B.rds")) |>
  select(ends_with("label"), ends_with("min"), ends_with("max"), ends_with("mid"))
```

```{r gpkg}
# ---- regA
regA_gpkg <- regA_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regA_gpkg, here::here("data/regionA_boxes.gpkg"), delete_layer = TRUE)

# ---- regB
regB_gpkg <- regB_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regB_gpkg, here::here("data/regionB_boxes.gpkg"), delete_layer = TRUE)
```

Lateral Subsurface Flow and location of selected regions 

```{r plot_location}

# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# ---- plot regB location ----
ggplot(data = world) +
  geom_sf(fill = "antiquewhite") +
  geom_point(data = regA_info, aes(x = xmid, y = ymid, color = strata_A_label), size = 3) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +
  labs(title = "Locations of Six Regions on the World Map",
       x = "Longitude",
       y = "Latitude",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "right")
```

Radiation and location of selected regions 

```{r}
# ---- plot regB location ----
ggplot(data = world) +
  geom_sf(fill = "antiquewhite") +
  geom_point(data = regB_info, aes(x = xmid, y = ymid, color = strata_B_label), size = 3) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +
  labs(title = "Locations of Six Regions on the World Map",
       x = "Longitude",
       y = "Latitude",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "right")
```

Additional plots

```{r}

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_A, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_B, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/04_re1_regions_selection.Rmd"))
```
