---
title: "Regions selection for topographic effect analysis"
author: "Ting Tan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
    fig_align: 'center'
bibliography: literatures.bib  
latex_engine: xelatex
csl: apa.csl
---

```{r SetUp, include=FALSE, eval=TRUE}

knitr::opts_chunk$set(
  comment = "#>", fig.width = 10, message = FALSE, warning = FALSE
)

library(terra)
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(DataExplorer)
library(RColorBrewer)
library(rnaturalearth)
library(sf)


# Load configuration and functions
source(here::here("config.R"))
source(here::here("R/plot_density_grid.R"))
source(here::here("R/extent_to_tile_ids.R"))
source(here::here("R/plot_vegh.R"))


if (!dir.exists(reg_sample_dir)) {
  dir.create(reg_sample_dir, recursive = TRUE)
}
```

---

## Task

Automatically identify regions where vegetation patterns are predominantly terrain-driven across different hydroclimatic and latitudinal backgrounds, in order to validate the Fan et al. (2019) framework.

## Implementation

```{r DataPre}
# Load resampled raster datasets (MI, TWI, fused)

mi_55km_r <- terra::rast(mi_55km_file) * 0.0001 # Multiply all values by 0.0001 to get original value
fused_55km_r <- terra::rast(fused_55km_file)
fbare_55km_r <- terra::rast(fbare_55km_file)
fwater_55km_r <- terra::rast(fwater_55km_file)
fsnow_55km_r <- terra::rast(fsnow_55km_file)
dem_sd_55km_r <- terra::rast(dem_sd_55km_path)
dem_rg_55km_r <- terra::rast(dem_rg_55km_path)
fpa_55km_r <- terra::rast(fpa_55km_path)
mat_55km_r <- terra::rast(mat_55km_file)
fveg_55km_r <- terra::rast(fveg_55km_path)

# Stack rasters with explicit layer names
stacked <- c(
  mi_55km_r,
  fused_55km_r,
  fbare_55km_r,
  fwater_55km_r,
  fsnow_55km_r,
  dem_sd_55km_r,
  dem_rg_55km_r,
  fpa_55km_r,
  mat_55km_r,
  fveg_55km_r
)

# Convert to data.frame 
df <- as.data.frame(stacked, xy = TRUE, na.rm = FALSE)
df <- df[complete.cases(df), ]

# easy to make mistake here!
colnames(df) <-  c("lon", "lat",
                   "mi", "fused", "fbare", "fwater", "fsnow",
                   "relief", "dem_rg", "fpa", "mat", "fveg")

# Clean up
rm(mi_55km_r, fused_55km_r, fbare_55km_r, fwater_55km_r, fsnow_55km_r, dem_sd_55km_r, dem_rg_55km_r, fpa_55km_r,  mat_55km_r, fveg_55km_r)
gc()


# load coast outline, vector data for plotting
coast <- rnaturalearth::ne_coastline(scale = 110, returnclass = "sf")
```

```{r}
# list all figures of input data 
files <- list.files(path = here::here("data/figures"), pattern = "^04_region_grid_.*_55km_map\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```


```{r DataOverview}
# Select columns for plot
cols <- c( "mi", "fused", "fbare","fwater","fsnow", "relief", "dem_rg", "fpa", "mat", "fveg")
# plot input data density
plot_density_grid(
  df,
  cols,
  nrow = 3,
  width = 14,
  height = 3,
  save_path = here::here("data/figures/04_region_selection_input_data_density.png")
)
```
---

### Binning Variables

To classify the global landscape into meaningful categories, we create bins for key drivers:

* **Moisture index (MI)**: split into five equal‐frequency bins (very dry → very humid).
* **Relief**: split into three equal‐frequency bins (flat → rugged).
* **Absolute latitude**: split into five equal‐frequency bins (low → high latitude).

From these, we define two sets of strata:

1. **Strata A** – combination of MI bin and relief bin (for lateral subsurface flow patterns).
2. **Strata B** – combination of latitude bin and relief bin (for insolation‐driven patterns).

Readable text labels (e.g., `"very_dry_flat_relief"`) are assigned for easier interpretation.

```{r Bin, echo = TRUE}
df_binned <- df |>

  # create bins
  mutate(
    # mi_bin = cut(mi, breaks = quantile(mi, probs = seq(0, 1, 1/5)), include.lowest = TRUE),
    mi_bin = cut(mi, breaks = c(0, 0.03, 0.2, 0.5, 0.65, 0.8, Inf), include.lowest = TRUE),
    relief_bin = cut(relief, breaks = quantile(relief, probs = seq(0, 1, 1/3)), include.lowest = TRUE),
    abs_lat = abs(lat),
    abs_lat_bin = cut(abs_lat, breaks = quantile(abs_lat, probs = seq(0, 1, 1/5)), include.lowest = TRUE)
  ) |>

  # Fig. 6 Down Valley drainage
  mutate(strata_A = interaction(mi_bin, relief_bin, drop = TRUE)) |>

  # Fig. 6 Aspect difference
  mutate(strata_B = interaction(abs_lat_bin, relief_bin, drop = TRUE)) |>

  # Assign readable labels to each bin
  mutate(
    mi_label = case_when(
      mi_bin == levels(mi_bin)[1] ~ "hyper_arid",
      mi_bin == levels(mi_bin)[2] ~ "semiarid",
      mi_bin == levels(mi_bin)[3] ~ "dry_subhumid",
      mi_bin == levels(mi_bin)[4] ~ "moist_subhumid",
      mi_bin == levels(mi_bin)[5] ~ "humid",
      mi_bin == levels(mi_bin)[6] ~ "perhumid"
    ),
    relief_label = case_when(
      relief_bin == levels(relief_bin)[1] ~ "flat_relief",
      relief_bin == levels(relief_bin)[2] ~ "moderate_relief",
      relief_bin == levels(relief_bin)[3] ~ "rugged_relief"
    ),
    lat_label = case_when(
      abs_lat_bin == levels(abs_lat_bin)[1] ~ "low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[2] ~ "relatively_low_latitude",
      abs_lat_bin == levels(abs_lat_bin)[3] ~ "mid_latitude",
      abs_lat_bin == levels(abs_lat_bin)[4] ~ "relatively_high_latitude",
      abs_lat_bin == levels(abs_lat_bin)[5] ~ "high_latitude"
    )
  ) |>

  # Combine text labels into new readable strata labels
  mutate(
    strata_A_label = paste0(mi_label, "_", relief_label),
    strata_B_label = paste0(lat_label, "_", relief_label)
  ) |>

  select(-mi_label, -relief_label, -lat_label)

# check data
table(df_binned$mi_bin)
table(df_binned$abs_lat_bin)
table(df_binned$relief_bin)

table(df_binned$strata_A_label)
table(df_binned$strata_B_label)
```

---

### Filtering Candidate Regions

#### For Lateral Subsurface Flow–Dominated Analysis (Strata A)

```{r FilterA, echo = TRUE}

df_filt_A <- df_binned |>
  
  filter(mi_bin %in% levels(mi_bin)[c(1, 2, 3, 4, 5, 6)],
         relief_bin %in% levels(relief_bin)[c(1, 3)],) |>
  
  filter(fused < 0.05) |>  # Exclude the region with high human activity 
  
  filter(fsnow < 0.1) |>  #  avoid high invalid or cold areas (permanent ice, and snow).
  
  filter(fwater < 0.3) |> # avoid regions with high invalid areas ( water bodies)
  
  # filter(fbare < 0.9) |>  # avoid regions with high invalid areas ( bare land)
  
  filter(fveg > 0.1) |> # avoid regions with very limited vegetation
  
  # filter(fpa > 0.3) |> # select regions with 30% protected areas

  filter(mat > 10)       # avoid cold regions for hydrology-specific analysis


# print the numbers of possible region
table(df_filt_A$strata_A_label)
```

#### For Insolation‐Dominated Analysis (Strata B)

We retain only regions with:

* Low human footprint and high protection status.
* Relief in flat or rugged categories.
* Latitude in very low, mid, or very high categories.

```{r FilterB, echo = TRUE}

df_filt_B <- df_binned |>
  
  filter(
    relief_bin %in% levels(relief_bin)[c(1, 3)],
    abs_lat_bin %in% levels(abs_lat_bin)[c(1, 2, 3, 4, 5)]
  ) |>
  
  filter(fused < 0.05) |>  # Exclude the region with high human activity
  
  filter(fsnow < 0.1) |>  #  avoid high invalid or cold areas (permanent ice, and snow).
  
  filter(fwater < 0.3) |> # avoid regions with high invalid areas ( water bodies)
  # 
  # filter(fbare < 0.9) |>  # avoid regions with high invalid areas ( bare land)
  
  filter(fveg > 0.1) |> # avoid regions with very limited vegetation, e.g. lined vegetation
  
  # filter(fpa > 0.3) |> # select regions with 30% protected areas
  
  filter(dem_rg < 500)   # avoid strong influence of elevation differences
  
# print the numbers of possible region
table(df_filt_B$strata_B_label)

```

---

### Sampling Representative Locations

```{r Sampling}
# Pick one region (coarse gridcell) representing all combination of bins.
# Should yield 6 samples

set.seed(2222)

# for Fig. 6 Down Valley drainage
df_samples_A <- df_filt_A |>
  group_by(strata_A) |>
  group_modify(~ slice_sample(.x, n = min(3, nrow(.x)))) |>
  mutate(
    xmin = round(lon - 0.25, 1),
    xmax = round(lon + 0.25, 1),
    ymin = round(lat - 0.25, 1),
    ymax = round(lat + 0.25, 1),
    xmid= (xmin + xmax)/2,
    ymid = (ymin + ymax)/2,
    sample_id = row_number()
  ) |>
  select(-starts_with("strata_B")) |>
  ungroup()

# for Fig. 6 Aspect difference
df_samples_B <- df_filt_B |>
  group_by(strata_B) |>
  group_modify(~ slice_sample(.x, n = min(3, nrow(.x)))) |>
  mutate(
    xmin = round(lon - 0.25, 1),
    xmax = round(lon + 0.25, 1),
    ymin = round(lat - 0.25, 1),
    ymax = round(lat + 0.25, 1),
    xmid= (xmin + xmax)/2,
    ymid = (ymin + ymax)/2,
    sample_id = row_number()
  ) |>
  select(-starts_with("strata_A")) |>
  ungroup()


# Save the two data frames as RDS files
saveRDS(df_samples_A, regA_sample_info_path)
saveRDS(df_samples_B, regB_sample_info_path)

```

---

### Export and Visualization

* Bounding boxes are saved as GeoPackage (`.gpkg`) files for GIS use.
* Locations are plotted on global maps, color‐coded by strata category.
* Additional maps show the global distribution of bins, filtered areas, and final selected regions.

```{r loadinfo}
# load region infos
regA_info <- readRDS(regA_sample_info_path)
regB_info <- readRDS(regB_sample_info_path) 

# ---- regA
regA_gpkg <- regA_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regA_gpkg,regA_sample_vect_path, delete_layer = TRUE)

# ---- regB
regB_gpkg <- regB_info |>
  rowwise() |>
  mutate(geometry = list(st_polygon(list(matrix(
    c(xmin, ymin,
      xmin, ymax,
      xmax, ymax,
      xmax, ymin,
      xmin, ymin),
    ncol = 2, byrow = TRUE))))) |>
  st_as_sf(crs = 4326)

# write gpkg document
st_write(regB_gpkg,regB_sample_vect_path, delete_layer = TRUE)
```

### region sample for lateral flow analysis

```{r plot_location}

# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# ---- plot regB location ----
ggplot(data = world) +
  geom_sf(fill = "antiquewhite") +
  geom_point(data = regA_info, aes(x = xmid, y = ymid, color = strata_A_label), size = 3) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +
  labs(title = "Sample regions for H~TWI",
       x = "Longitude",
       y = "Latitude",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# regA_info
# regB_info
plot_vegh_by_reg <- function(reg_row,
                              text_size = 12,
                              fig_width = 14,
                              fig_height = 18) {

    # --- reg info ---
    reg_id <- paste0(reg_row$strata_A_label, "_", reg_row$sample_id)
    reg_extent <- terra::ext(reg_row$xmin, reg_row$xmax, reg_row$ymin, reg_row$ymax)
    reg_xmid <- reg_row$lon
    reg_ymid <- reg_row$lat

    # --- Vegetation Height Raster ---
    vegh_rc <- terra::rast(vegh_450m_mosaic_path) |> terra::crop(reg_extent)
    vegh_tmp_path <- file.path(tempdir(), paste0("tile_", reg_id, "_vegh_450m.nc"))
    terra::writeCDF(vegh_rc, vegh_tmp_path, varnames = "vegh",overwrite = TRUE)
    rm(vegh_rc); gc()
    vegh_rc <- terra::rast(vegh_tmp_path)
    names(vegh_rc) <- "vegh"
    
    p_vegh <- plot_vegh(vegh_rc, extent = reg_extent, title_text = reg_id, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    
    return(p_vegh)
    
    rm(vegh_rc);gc()

}

for (i in seq_len(nrow(regA_info))) {
  p <- plot_vegh_by_reg(regA_info[i, ])
  print(p)
}
```


### region sample for Radiation analysis
```{r}
# ---- plot regB location ----
ggplot(data = world) +
  geom_sf(fill = "antiquewhite") +
  geom_point(data = regB_info, aes(x = xmid, y = ymid, color = strata_B_label), size = 3) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +
  labs(title = "Sample regions for H~Rin",
       x = "Longitude",
       y = "Latitude",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# regA_info
# regB_info
plot_vegh_by_reg <- function(reg_row,
                              text_size = 12,
                              fig_width = 14,
                              fig_height = 18) {

    # --- reg info ---
    reg_id <- paste0(reg_row$strata_B_label, "_", reg_row$sample_id)
    reg_extent <- terra::ext(reg_row$xmin, reg_row$xmax, reg_row$ymin, reg_row$ymax)
    reg_xmid <- reg_row$lon
    reg_ymid <- reg_row$lat

    # --- Vegetation Height Raster ---
    vegh_rc <- terra::rast(vegh_450m_mosaic_path) |> terra::crop(reg_extent)
    vegh_tmp_path <- file.path(tempdir(), paste0("tile_", reg_id, "_vegh_450m.nc"))
    terra::writeCDF(vegh_rc, vegh_tmp_path, varnames = "vegh",overwrite = TRUE)
    rm(vegh_rc); gc()
    vegh_rc <- terra::rast(vegh_tmp_path)
    names(vegh_rc) <- "vegh"
    
    p_vegh <- plot_vegh(vegh_rc, extent = reg_extent, title_text = reg_id, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    
    return(p_vegh)
    
    rm(vegh_rc);gc()

}

for (i in seq_len(nrow(regB_info))) {
  p <- plot_vegh_by_reg(regB_info[i, ])
  print(p)
}
```

Additional plots

```{r}

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_binned, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_A, aes(x = lon, y = lat, fill = strata_A)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()

ggplot(df_filt_B, aes(x = lon, y = lat, fill = strata_B)) +
  geom_tile() +
  geom_sf(data = coast, colour = 'black', linewidth = 0.1, inherit.aes = FALSE) +
  theme_bw() +
  coord_sf()
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/04_re1_regions_selection.Rmd"))
```
