---
title: "r_H_R Calculation on defined regions"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# ------Load required libraries-------------------------------------------------

library(terra)     # For handling raster data
library(furrr)
library(dplyr)
library(ggmap)
library(tidyterra)
library(doParallel)    # inner parallelism
library(foreach)

library(khroma)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
library(rnaturalearth)
library(sf)

# ------Load configuration and helper functions---------------------------------

# Automatically select configuration file
hostname <- trimws(tolower(system("hostname", intern = TRUE)))
if (hostname == "dash") {
  message("💻 Detected Worksation: dash → using config.R")
  source(here::here("config.R"))
  workers = 16 # don't do it on worksation
} else {
  message("🖥️ Detected HPC environment (", hostname, ") → using config_ubelix.R")
  source(here::here("config_ubelix.R"))
  workers = 100
}
source(here::here("R/create_spatial_windows.R"))
source(here::here("R/calculate_correlation_bywin.R"))
source(here::here("R/mosaic_tiles.R"))
source(here::here("R/plot_dem.R"))
source(here::here("R/plot_aspect.R"))
source(here::here("R/plot_slope.R"))
source(here::here("R/plot_vegh.R"))
source(here::here("R/plot_r_H_R.R"))
source(here::here("R/plot_sw_in.R"))
source(here::here("R/plot_rin.R"))
source(here::here("R/plot_hex_scatter.R"))
source(here::here("R/plot_single_sample_location.R"))
source(here::here("R/plot_google_img.R"))
# source(here::here("R/plot_cci_land_cover.R"))
source(here::here("R/plot_fused.R"))
source(here::here("R/plot_cor_pval.R"))
source(here::here("R/plot_cor_twi_vegh.R"))

source(here::here("R/raster_preprocess_save.R"))
source(here::here("R/aggregate_topography.R"))
source(here::here("R/extent_to_tile_ids.R"))

source(here::here("R/helpers.R")) # SPLASH
source(here::here("R/calc_sw_in.R")) # SPLASH
```


---

## Task

To verify the correctness of the split-window correlation analysis, all input and intermediate datasets should be visualized to ensure the accuracy of the workflow.

We manually selected several well-known regions, including those mentioned by Fan et al. (2019), for this validation.
The evaluation will be conducted at two spatial scales:

1. Coarse resolution (5 km): Pearson’s r will be computed using the radiation index and vegetation height aggregated to 450 m.
2. Fine resolution (500 m): Pearson’s r will be calculated using the radiation index and vegetation height data at 30 m resolution.

This setup allows us to assess whether the analysis performs consistently across different spatial resolutions and datasets.

---

## Define regions

```{r}

tiles_info <- data.frame(
  tile_id = c(
    "Aletsch_glacier",
    "3a_arid_western_Texas",
    "3b_seasonal_arid_California",
    "3c_seasonal_arid_Idaho",
    "3d_winter_snowpacks_Mount_Cramer",
    "3e_high_latitude_southwest_Yukon",
    "3f_boreal_forest_Alaska"
  ),
  ymin = c(46.8, 31.5, 37.5, 43, 43.8, 61.5, 64),
  ymax = c(47, 32.5, 38.5, 44, 44.1, 62.5, 65),
  xmin = c(7.8, -105, -120.5, -115.8, -115.5, -138.5, -148),
  xmax = c(8, -104.5, -119.5, -114.8, -114.5, -137.5, -147)
)

```

```{r, eval=FALSE}
regB_row <- tiles_info[1, ]
output_dir = tempdir()
text_size = 12
fig_width = 14
fig_height = 18

```

---

## Global analysis on validation regions

```{r res1, eval=FALSE}
process_r_H_R_5km <- function(tile_row, output_dir = tempdir(),
                                text_size = 12, fig_width = 14, fig_height = 18) {

  tryCatch({


    # --- Tile info ---
    tile_id <- tile_row$tile_id
    tile_extent <- terra::ext(tile_row$xmin, tile_row$xmax, tile_row$ymin, tile_row$ymax)
    tile_xmid <- (tile_row$xmin + tile_row$xmax)/2
    tile_ymid <- (tile_row$ymin + tile_row$ymax)/2

    tictoc::tic(paste0("Processing tile: ", tile_id))
    t0 <- Sys.time()

    # --- TWI Raster ---
    twi_rc <- terra::rast(twi_450m_mosaic_clean_path) |> terra::crop(tile_extent)
    twi_tmp_path <- file.path(tempdir(), paste0("tile_", tile_id, "_twi_450m.nc"))
    terra::writeCDF(twi_rc, twi_tmp_path, varnames = "twi", overwrite = TRUE)
    rm(twi_rc); gc()
    twi_rc <- terra::rast(twi_tmp_path)
    names(twi_rc) <- "twi"
    message("Saved temporary TWI raster: ", twi_tmp_path)


    # --- Vegetation Height Raster ---
    vegh_rc <- terra::rast(vegh_450m_mosaic_path) |> terra::crop(tile_extent)
    vegh_tmp_path <- file.path(tempdir(), paste0("tile_", tile_id, "_vegh_450m.nc"))
    terra::writeCDF(vegh_rc, vegh_tmp_path, varnames = "vegh",overwrite = TRUE)
    rm(vegh_rc); gc()
    vegh_rc <- terra::rast(vegh_tmp_path)
    names(vegh_rc) <- "vegh"
    message("Saved temporary vegetation height raster: ", vegh_tmp_path)

    # --- Radiation index of terrain effect ---
    rin_rc <- terra::rast(sw_in_terrain_effect_450m_path) |> terra::crop(tile_extent)
    terrain_tmp_path <- file.path(tempdir(), paste0("tile_", tile_id, "_terrain_effect_450m.nc"))
    terra::writeCDF(rin_rc, terrain_tmp_path, varnames = "rin", overwrite = TRUE)
    rm(rin_rc); gc()
    rin_rc <- terra::rast(terrain_tmp_path)
    names(rin_rc) <- "rin"
    message("Saved temporary radiation index raster: ", terrain_tmp_path)

    # # --- Elevation and radiation Raster ---
    # dem_rc <- extent_to_tile_ids(tile_extent, tile_size = 1, return_raster = TRUE,
    #                              source = "copernicus_dem_30m", tiles_dir = dem_30m_copernicus_dir)
    #
    # # Aggregate DEM and calculate slope/aspect
    # aligned <- aggregate_topography(
    #   dem_rc,
    #   res_tar = NULL,
    #   target = twi_rc,
    #   if_resample = TRUE
    # )
    # rm(dem_rc); gc()

    # --- Stack and correlation ---
    stacked <- c(rin_rc, vegh_rc)
    df_win <- create_spatial_windows(stacked, value_vars = c("rin", "vegh"), dwin = 0.05)
    df_cor <- calculate_correlation_bywin(df_win, x = "rin", y = "vegh")

    # --- Save correlation ---
    cor_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "correlation")], type="xyz", crs="EPSG:4326")
    names(cor_r) <- "r_H_R"
    cor_nc_path <- file.path(output_dir, paste0("tile_", tile_id, "_r_H_R_5km_map.nc"))
    terra::writeCDF(cor_r, cor_nc_path, overwrite = TRUE)
    message("Saved: ", cor_nc_path)

    # --- Save p-value ---
    pval_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "cor_pval")], type="xyz", crs="EPSG:4326")
    names(pval_r) <- "pval_r_H_R"
    pval_nc_path <- file.path(output_dir, paste0("tile_", tile_id, "_r_H_R_5km_pval.nc"))
    terra::writeCDF(pval_r, pval_nc_path, overwrite = TRUE)
    message("Saved: ", pval_nc_path)

    # ---- Generate plots ----
    p_dem <- plot_dem(dem_450m_mosaic_path, extent = tile_extent, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_slope <- plot_slope(slope_450m_mosaic_path, extent = tile_extent, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_aspect <- plot_aspect(aspect_450m_mosaic_path, extent = tile_extent, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_vegh <- plot_vegh(vegh_rc, extent = tile_extent, text_size = text_size)  + ggplot2::theme(aspect.ratio = 1)
    p_rin <- plot_rin(rin_rc, extent = tile_extent, text_size = text_size)  + ggplot2::theme(aspect.ratio = 1)
    p_r_H_R <- plot_r_H_R(cor_r, extent = tile_extent, title_text = "5km: Pearson's r (H~R)", text_size = text_size)  + ggplot2::theme(aspect.ratio = 1)
    p_p <- plot_cor_pval(pval_r, extent = tile_extent,  title_text = "5km: Pearson's p value (H~R)", text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_r_H_TWI <- plot_cor_twi_vegh(cor_twi_vegh_mosaic_file, extent = tile_extent, title_text = "5km: Pearson's r (H~TWI)", text_size = text_size)  + ggplot2::theme(aspect.ratio = 1)
    p_google <- plot_google_img(extent = tile_extent) + ggplot2::theme(aspect.ratio = 1)
    p_fused <- plot_fused(fused_5km_file, extent = tile_extent, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_scatter <- plot_hex_scatter(df_win, x_var = "rin", y_var = "vegh", x_text = "Radiation index", y_text = "Vegetation height (m)", text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_location <- plot_single_sample_location(tile_xmid, tile_ymid, tile_id, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    # ---- Combine plots ----
    final_plot <- (
      (p_dem + p_slope + p_aspect) /
        (p_rin + p_vegh + p_fused) /
        (p_r_H_R + p_p + p_r_H_TWI )  /
        (p_google + p_location + p_scatter)) +
      plot_annotation(title = tile_id) +
      plot_layout(heights = c( 1, 1, 1, 1))

    # ---- Save plot ----
    out_file <- here::here(file.path(paste0("data/figures/05_validate_", tile_id, "_win_5km_plots.png")))
    ggsave(filename = out_file, plot = final_plot, width = fig_width, height = fig_height, dpi = 600)

    # ---- Memory cleanup ----
    # list all objects need to be remove
    rm(
      p_dem, p_slope, p_aspect, p_p,
      twi_rc, vegh_rc, rin_rc,
      stacked, df_win, df_cor, cor_r,pval_r,
      p_vegh, p_rin, p_r_H_R,
      p_r_H_TWI, p_google, p_fused, p_scatter, p_location,
      final_plot)
    gc(verbose = FALSE)

    # --- Print proccessed time ---
    elapsed_mins <- difftime(Sys.time(), t0, units = "mins")
    message(sprintf("Region %s completed [%.1f mins]", tile_id, elapsed_mins))
    tictoc::toc()

    return(TRUE)


  }, error = function(e) {
    elapsed_mins <- difftime(Sys.time(), t0, units = "mins")
    message(sprintf("❌ Tile %s failed after %.1f mins: %s", tile_row$tile_id, elapsed_mins, e$message))
    return(FALSE)
  })
}


for (i in seq_len(nrow(tiles_info))) {
  process_r_H_R_5km(tiles_info[i, ])
}
```

---


## Regional Analysis on validation regions

```{r res2, eval=TRUE}
process_regB_500m <- function(regB_row,
                              output_dir = tempdir(),
                              text_size = 12,
                              fig_width = 14,
                              fig_height = 18) {
  tryCatch({

    # --- reg info ---

    regB_id <- regB_row$tile_id
    regB_extent <- terra::ext(regB_row$xmin, regB_row$xmax, regB_row$ymin, regB_row$ymax)
    regB_xmid <- (regB_row$xmin +regB_row$xmax) / 2
    regB_ymid <- (regB_row$ymin +regB_row$ymax) / 2

    # Start Processing
    tictoc::tic(paste0("Processing tile: ", regB_id))
    t0 <- Sys.time()

    # --- TWI Raster ---
    twi_rc <- terra::rast(twi_30m_path) |> terra::crop(regB_extent)
    twi_nc_path <- file.path(output_dir, paste0("regB_tile_", regB_id, "_twi_450m.nc"))
    terra::writeRaster(twi_rc, twi_nc_path, overwrite = TRUE, 
                   wopt = list(gdal = c("COMPRESS=LZW")))
    rm(twi_rc); gc()
    twi_rc <- terra::rast(twi_nc_path)
    names(twi_rc) <- "twi"
    message("Saved: ", twi_nc_path)

    # --- Vegetation Height Raster ---
    vegh_rc <- extent_to_tile_ids(regB_extent, tile_size = 3, return_raster = TRUE,
                                  source = "lang_vegh_10m", tiles_dir = vegh_10m_tiles_dir)

    # Vegetation Height Raster: crop and store in temporary file
    tmp_vegh <- tempfile(fileext = ".tif")
    terra::writeRaster(vegh_rc, tmp_vegh, overwrite = TRUE,
                   wopt = list(gdal = c("COMPRESS=LZW")))
    rm(vegh_rc); gc()
    vegh_rc <- terra::rast(tmp_vegh)
    names(vegh_rc) <- "vegh"

    # Set 0 as NA value (0m canopy height represents not vegetated or water according to Lang et al. (2019))
    # Aggregate and resample using TWI data from Ho et al. (2025)
    vegh_rc <- raster_preprocess_save(
      input = vegh_rc,
      target = twi_rc,
      na_value = 0,
      fun = mean,
      varname = "vegh",
      if_aggregate = TRUE,
      if_resample = TRUE,
      if_mask = TRUE,
      if_return_raster = TRUE
    )

    # --- Elevation and radiation Raster ---
    dem_rc <- extent_to_tile_ids(regB_extent, tile_size = 1, return_raster = TRUE,
                                 source = "copernicus_dem_30m", tiles_dir = dem_30m_copernicus_dir)

    # Aggregate DEM and calculate slope/aspect
    aligned <- aggregate_topography(
      dem_rc,
      res_tar = NULL,
      target = twi_rc,
      if_resample = TRUE
    )

    # --- Data frame Prepare ---

    # Extract + join
    df <- as.data.frame(aligned[["dem"]], xy = TRUE) |>
      left_join(as.data.frame(aligned[["slope"]], xy = TRUE), by = c("x", "y")) |>
      left_join(as.data.frame(aligned[["aspect"]], xy = TRUE), by = c("x", "y")) |>
      tibble::as_tibble() |>
      drop_na()
    names(df) <- c("lon", "lat", "dem", "slope", "aspect")

    if (nrow(df) == 0) {
      warning(sprintf("No valid cells after drop_na for %s", file))
      return(FALSE)
    }

    # --- Inner Paralell ---

    # clean old cluster
    try({
      if (exists("cl")) stopCluster(cl)
      closeAllConnections()
    }, silent = TRUE)

    # Inner parallelism
    num_cores <- wokers
    cl <- makeCluster(num_cores)
    registerDoParallel(cl)

    # Chunk Processing
    chunk_size <- 5000  # rows per chunk, adjust based on memory
    chunks <- split(df, ceiling(seq_len(nrow(df)) / chunk_size))

    # Parallel Calculation for Each Chunk - Direct assignment
    df_calc <- foreach(
      chunk = chunks,
      .combine = bind_rows,
      .packages = c("dplyr"),
      .export   = c(
        "calc_sw_in_daily",
        "calc_sw_in",
        "julian_day",
        "berger_tls",
        "dcos",
        "dsin"
      )
    ) %dopar% {
      # Calculate sw_in_uneven and sw_in_flat for entire chunk
      sw_in_uneven <- calc_sw_in(chunk$lat, chunk$slope, chunk$aspect, year = 2020)
      sw_in_flat <- calc_sw_in(chunk$lat, rep(0, nrow(chunk)), rep(0, nrow(chunk)), year = 2020)

      # Combine results back to dataframe
      chunk |>
        mutate(
          sw_in_uneven = sw_in_uneven,
          sw_in_flat = sw_in_flat,
          rin = sw_in_uneven / sw_in_flat
        )

    }

    # Stop cluster
    stopCluster(cl)
    registerDoSEQ()

    # --- Build rasters ---
    crs_out <- terra::crs(aligned[["dem"]])
    sw_in_uneven <- terra::rast(df_calc[, c("lon", "lat", "sw_in_uneven")], type = "xyz", crs = crs_out)
    sw_in_flat <- terra::rast(df_calc[, c("lon", "lat", "sw_in_flat")], type = "xyz", crs = crs_out)
    rin <- terra::rast(df_calc[, c("lon", "lat", "rin")], type = "xyz", crs = crs_out)
    rin <- terra::resample(rin, twi_rc, method = "near")

    # --- Stack and Correalation---
    stacked <- c(rin, vegh_rc)
    names(stacked) <- c("rin", "vegh")
    df_win <- create_spatial_windows(stacked, value_vars = c("rin", "vegh"), dwin = 0.005)
    df_cor <- calculate_correlation_bywin(df_win, x = "rin", y = "vegh")

    # --- Save correlation ---
    cor_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "correlation")], type = "xyz", crs = "EPSG:4326")
    names(cor_r) <- "r_H_R"

    # --- Save p-value ---
    pval_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "cor_pval")], type = "xyz", crs = "EPSG:4326")
    names(pval_r) <- "pval_r_H_R"

    # ---- Generate plots ----
    p_dem <- plot_dem(aligned[["dem"]], extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_slope <- plot_slope(aligned[["slope"]], extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_aspect <- plot_aspect(aligned[["aspect"]], extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_vegh <- plot_vegh(vegh_rc, extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_rin <- plot_rin(rin, extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_r_H_R <- plot_r_H_R(cor_r, extent = regB_extent, title_text = "500m: Pearson's r (H ~ R)", text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_r_H_R2 <- plot_r_H_R(r_H_R_5km_path, extent = regB_extent, title_text = "5km: Pearson's r (H ~ R)", text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_google <- plot_google_img(extent = regB_extent) + ggplot2::theme(aspect.ratio = 1)
    p_fused <- plot_fused(fused_5km_file, extent = regB_extent, text_size = text_size, x_step = 0.5, y_step = 0.5) + ggplot2::theme(aspect.ratio = 1)
    p_scatter <- plot_hex_scatter(df_win, x_var = "rin", y_var = "vegh", x_text = "Radiation index", y_text = "Vegetation height (m)", text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    p_location <- plot_single_sample_location(regB_xmid, regB_ymid, regB_id, text_size = text_size) + ggplot2::theme(aspect.ratio = 1)
    # ---- Combine plots ----
    final_plot <- ((p_dem + p_slope + p_aspect) /
                     (p_r_H_R + p_vegh + p_rin) /
                     (p_r_H_R2 + p_location + p_scatter) /
                     (p_google + p_fused)) +
      plot_annotation(title = regB_id) +
      plot_layout(heights = c(1, 1, 1, 1))

    # ---- Save plot ----
    out_file <- here::here(file.path(paste0("data/figures/05_validate_", regB_id, "_win_500m_plots.png")))
    ggsave(filename = out_file, plot = final_plot, width = fig_width, height = fig_height, dpi = 600)

    # ---- Memory cleanup ----
    # list all objects need to be remove
    rm(twi_rc, vegh_rc, dem_rc, aligned,
       df, chunks, df_calc, stacked, df_win, df_cor, cor_r,pval_r,
       p_dem, p_slope, p_aspect, p_vegh, p_rin, p_r_H_R,
       p_r_H_R2, p_google, p_fused, p_scatter, p_location,
       final_plot)
    gc(verbose = FALSE)

    # --- Print proccessed time ---
    elapsed_mins <- difftime(Sys.time(), t0, units = "mins")
    message(sprintf("Region %s completed [%.1f mins]", regB_id, elapsed_mins))
    tictoc::toc()

    return(TRUE)

  }, error = function(e) {
    regB_id <- paste0(regB_row$strata_B_label, "_", regB_row$sample_id)
    elapsed_mins <- difftime(Sys.time(), t0, units = "mins")
    message(sprintf("❌ Tile %s failed after %.1f mins: %s", regB_id, elapsed_mins, e$message))
    return(FALSE)
  })
}

for (i in seq_len(nrow(tiles_info))) {
  process_regB_500m(tiles_info[i, ])
}
```

---
### Aletsch Glacier
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_Aletsch.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3a_arid_western_Texas
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3a_arid_western_Texas.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3b_seasonal_arid_California
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3b_seasonal_arid_California.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3c_seasonal_arid_Idaho
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3c_seasonal_arid_Idaho.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3d_winter_snowpacks_Mount_Cramer
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3d_winter_snowpacks_Mount_Cramer.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3e_high_latitude_southwest_Yukon
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3e_high_latitude_southwest_Yukon.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### 3f_boreal_forest_Alaska
5km vs 500m
```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "05_validate_3f_boreal_forest_Alaska.*\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

###  5km vs 500m
```{r}
# # list all figures
# # files <- list.files(path = here::here("data/figures"), pattern = "05_validate.*5km_.*\\.png$", full.names = TRUE)
# 
# # Print 
# cat("Found", length(files), "figures:\n")
# 
# # print
# knitr::include_graphics(files)
```

```{r}
# 
# # list all figures
# files <- list.files(path = here::here("data/figures"), pattern = "05_validate_.*500m_.*\\.png$", full.names = TRUE)
# 
# # Print 
# cat("Found", length(files), "figures:\n")
# 
# # print
# knitr::include_graphics(files)
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/05_re3.Rmd"))
```
