---
title: "Comparison of Incident Radiation Estimates from SPLASH 2.0 and meteoland"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(terra)       # Raster handling
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(lubridate)
library(meteoland)   # Radiation functions

source(here::here("config.R"))
source(here::here("R/process_raster.R"))
source(here::here("R/batch_process_rasters.R"))

source(here::here("R/helpers.R"))
source(here::here("R/calc_sw_in.R")) # SPLASH
source(here::here("R/compute_annual_radiation.R")) # meteoland

source(here::here("R/create_spatial_windows.R"))
source(here::here("R/calculate_correlation_bywin.R"))
```


# Introduction

Topography strongly influences incoming solar radiation, which in turn drives systematic variation in vegetation productivity along topographic gradients.

To test the accuracy and applicability of code adapted from SPLASH 2.0 (Sandoval et al.), we compared its estimates with those from the published R package {meteoland}.

This comparison allows us to evaluate the feasibility of SPLASH-based radiation calculations relative to an established implementation.

---

# Data Pre

## Read DEM and calculate slope & aspect 

```{r}
# Read DEM (GeoTIFF) and target raster 
dem <- rast("/data_2/scratch/ting/veg_topo_data/data_raw/copernicus_dem_30m/copernicus_dem_30m/Copernicus_DSM_COG_10_N67_00_W155_00_DEM/Copernicus_DSM_COG_10_N67_00_W155_00_DEM.tif")

# reduce extent to example region
# extent <- ext(-154.5,-154.25, 67.25, 67.5)
extent <- ext(-154.5,-154, 67.25, 67.75)
dem_crop <- crop(dem, extent)

# Calculate slope and aspect (in degrees)
slope <- terrain(dem_crop, v="slope", unit="degrees")
aspect <- terrain(dem_crop, v="aspect", unit="degrees")
```

Load vegetation Data, the aggregation seems to be not correct due to the NA handling.

```{r}
# load and crop vegh to calculate r(R~H) 
vegh <- rast(vegh_450m_mosaic_path)
vegh_crop <- terra::crop(vegh, terra::ext(dem_crop)) 

plot(vegh)
```

---

## Aggregate to 15 arcsec resolution (as TWI data from Marthew et. al)

```{r}
# target layer for aggregation
# Crop TWI to DEM extent first, otherwise resampling DEM to TWI would expand it to global extent
twi <- rast(twi_450m_mosaic_clean_path)
twi_crop <- terra::crop(twi, terra::ext(dem_crop)) # need to crop it first, otherwise the output extent of resampled one is global.

# Aggregate to 450m resolution as TWI data from Marthew et. al.
aligned <- batch_process_rasters(list(dem = dem_crop, slope = slope, aspect = aspect), target = twi_crop)
tictoc::toc()

```

---

## Extract data from raster

```{r}
df <- as.data.frame(aligned[["dem"]], xy = TRUE) |> 
  dplyr::left_join(
    as.data.frame(aligned[["slope"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  dplyr::left_join(
    as.data.frame(aligned[["aspect"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  dplyr::left_join(
    as.data.frame(vegh_crop, xy = TRUE),
    by = c("x", "y")
  ) |>
  as_tibble() |> 
  drop_na()

names(df) = c("lon", "lat", "dem", "slope" ,"aspect", "vegh")
```

---

## Input Data Visualization

```{r,fig.width=8,fig.height=12}

p_dem <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["dem"]]) +
  scale_fill_viridis_c() +
  labs(title = "Elevation (m)", fill = "Elevation") +
  theme_bw()

p_slope <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["slope"]]) +
  scale_fill_viridis_c() +
  labs(title = "Slope (°)", fill = "Slope") +
  theme_bw()

# Define a smooth circular color palette
my_colors <- c("navy", "#88ccee", "white", "#ff8888", "red", "#ff8888", "white", "#88ccee", "navy")
my_values <- seq(0, 360, length.out = length(my_colors))/360  # normalize 0-360 to 0-1

p_aspect <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["aspect"]]) +
  scale_fill_gradientn(
    colors = my_colors,
    values = my_values,
    limits = c(0, 360),
    oob = scales::squish
  ) +
  labs(title = "Aspect (°)", fill = "Aspect") +
  theme_bw()

p_vegh <- ggplot() +
  tidyterra::geom_spatraster(data = vegh_crop) +
  scale_fill_viridis_c() +
  labs(title = "vegetation Heigh (m)", fill = "vegh") +
  theme_bw()

# print
p_dem + p_vegh + p_slope + p_aspect
```

# Comparision

---

## SPLASH 2.0

```{r}
tictoc::tic()
df$splash_rad <- sapply(1:nrow(df), function(i){
  calc_sw_in(
    lat = df$lat[i],
    slope = df$slope[i],
    aspect = df$aspect[i],
    year = 2020
  )  / 1e6
})
tictoc::toc()
```

---

## `meteoland::radiation_potentialRadiation`

```{r}
tictoc::tic()
df$meteoland_rad <- sapply(1:nrow(df), function(i){
  compute_annual_radiation(
    lat_deg = df$lat[i],
    slope_deg = df$slope[i],
    aspect_deg = df$aspect[i],
    year = 2020
  )
})
tictoc::toc()
```


## Scatterplot Comparison

```{r,fig.width=6,fig.height=4}
ggplot(df, aes(x = splash_rad, y = meteoland_rad)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    x = "SPLASH Radiation",
    y = "Meteoland Radiation",
    title = "Comparison of SPLASH vs Meteoland Radiation"
  ) +
  theme_bw()
```

## Density Distributions

```{r,fig.width=6,fig.height=4}
df_long <- df %>%
  pivot_longer(cols = c(splash_rad, meteoland_rad),
               names_to = "source",
               values_to = "radiation")

ggplot(df_long, aes(x = radiation, fill = source, color = source)) +
  geom_density(alpha = 0.4) +
  labs(x = "Radiation", y = "Density", title = "Density distribution of SPLASH vs Meteoland") +
  theme_minimal()
```

## Spatial Visualization

```{r, fig.width=8,fig.height=12}
p_splash <- ggplot(df, aes(x = lon, y = lat, fill = splash_rad)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "SPLASH Radiation",
    title = "Spatial Distribution of SPLASH Radiation"
  ) +
  coord_fixed() +
  theme_bw()

p_meteoland <- ggplot(df, aes(x = lon, y = lat, fill = meteoland_rad)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "meteoland Radiation",
    title = "Spatial Distribution of meteoland Radiation"
  ) +
  coord_fixed() +
  theme_bw()

# print
p_dem + p_vegh + p_splash + p_meteoland
```

# Calculate r(R~H)

In high-latitude regions, vegetation on the south slope, which receives stronger incident solar radiation, shows lower production efficiency than on the north slope. This is contrary to the framework proposed by Fan et al. (2019).


```{r}
gc()

dwin <- 0.05  # ~500m window at equator
df_win <- create_spatial_windows(df, 
                                 coord_vars = c("lon", "lat"),
                                 value_vars = c( "dem", "slope", "aspect", "splash_rad", "meteoland_rad", "vegh"),
                                 dwin = dwin)  
colnames(df_win)

df_cor <- calculate_correlation_bywin(df_win, 
                                      x = "splash_rad",
                                      y = "vegh")            

colnames(df_cor)
```


```{r}
vmin <- min(df_cor$correlation, na.rm = TRUE)
vmax <- max(df_cor$correlation, na.rm = TRUE)

p_cor <- ggplot(df_cor, aes(x = lon_mid, y = lat_mid, fill = correlation)) +
  geom_tile() +
  scico::scale_fill_scico(
      palette = "bam",
      limits = c(vmin, vmax),
      midpoint = 0,
      name = expression(r[R*","*H]),
      na.value = NA
    ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "r(R~H)",
    title = "r(R~H)"
  ) +
  coord_fixed() +
  theme_bw()


p_meteoland + p_cor
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/05_re1_compare_radiation_methods.Rmd"))
```
