---
title: "Checking the complement and numes of calculated incident solar radiation document"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

## Task

We are performing a parallelized calculation of **surface solar radiation** over **26,450 1×1 degree tiles**. During processing, some outputs may encounter errors, resulting in **broken or incomplete files**.

The goal is to:

1. **Verify the outputs** by checking the number of generated files.
2. **Assess their quality** to detect incomplete or corrupted results.
3. **Identify missing or broken tiles** and print their corresponding **tile IDs** for further action.

---

```{r setup, include=FALSE}
library(terra)
library(fs)
library(ncdf4)
# source(here::here("config.R"))
```

```{r}
# ---- Data Source Configuration ----
dem_30m_copernicus_dir <- "/storage/scratch/giub_geco/tting/copernicus_dem_30m/copernicus_dem_30m"
sw_in_450m_tile_dir <- "/storage/scratch/giub_geco/tting/global_sw_in_450m/1_1_deg_tiles"
```

```{r}
# Read input DEM files
dem_files_all <- fs::dir_ls(
  path = dem_30m_copernicus_dir,
  glob = "*_DEM.tif",
  recurse = TRUE
)

message(sprintf("Found %d DEM tiles", length(dem_files_all)))

```

```{r}
# Get processed SW_IN files
sw_in_uneven_files <- list.files(
  path = sw_in_450m_tile_dir,
  pattern = "*_to_sw_in_uneven_450m\\.nc$",
  recursive = TRUE,
  full.names = TRUE
)

sw_in_flat_files <- list.files(
  path = sw_in_450m_tile_dir,
  pattern = "*_to_sw_in_flat_450m\\.nc$",
  recursive = TRUE,
  full.names = TRUE
)

# Extract base names for comparison
dem_base_names <- gsub("\\.tif$", "", basename(dem_files_all))
sw_in_uneven_base_names <- gsub("_to_sw_in_uneven_450m\\.nc$", "", basename(sw_in_uneven_files))
sw_in_flat_base_names <- gsub("_to_sw_in_flat_450m\\.nc$", "", basename(sw_in_flat_files))

# Find missing files
missing_sw_in_uneven <- setdiff(dem_base_names, sw_in_uneven_base_names)
missing_sw_in_flat <- setdiff(dem_base_names, sw_in_flat_base_names)

message(sprintf("Missing uneven SW_IN files: %d", length(missing_sw_in_uneven)))
message(sprintf("Missing flat SW_IN files: %d", length(missing_sw_in_flat)))

```

```{r}
# Check a single tile (simplified)
check_tile <- function(tile_index) {
  if (tile_index > length(dem_files_all)) {
    message("Tile index out of range")
    return(NULL)
  }

  dem_file <- dem_files_all[tile_index]
  dem_base_name <- gsub("\\.tif$", "", basename(dem_file))

  sw_in_uneven <- file.path(sw_in_450m_tile_dir, paste0(dem_base_name, "_to_sw_in_uneven_450m.nc"))
  sw_in_flat <- file.path(sw_in_450m_tile_dir, paste0(dem_base_name, "_to_sw_in_flat_450m.nc"))

  # Check if output files exist
  if (!file.exists(sw_in_uneven) || !file.exists(sw_in_flat)) {
    message(sprintf("Output files missing for: %s", dem_base_name))
    return(NULL)
  }

  # Read and return rasters
  dem_raster <- rast(dem_file)
  uneven_raster <- rast(sw_in_uneven)
  flat_raster <- rast(sw_in_flat)

  return(list(dem = dem_raster, uneven = uneven_raster, flat = flat_raster))
}
```

```{r}
# Example: Check tile 1199
tile_data <- check_tile(199)
if (!is.null(tile_data)) {
  par(mfrow = c(1, 3))
  plot(tile_data$dem, main = "DEM")
  plot(tile_data$uneven, main = "Uneven SW_IN")
  plot(tile_data$flat, main = "Flat SW_IN")
}
```


```{r}
# Check if a single NetCDF file is OK
is_nc_broken <- function(file_path) {
  # Check header with ncdf4
  nc_ok <- !inherits(try(nc_open(file_path), silent = TRUE), "try-error")
  if (nc_ok) nc_close(nc_open(file_path))
  
  # Check if terra can read it
  terra_ok <- !inherits(try(rast(file_path), silent = TRUE), "try-error")
  
  # Return TRUE if broken, FALSE if OK
  return(!(nc_ok & terra_ok))
}

# Traverse directory and return only broken file names
get_broken_nc_files <- function(dir_path) {
  nc_files <- list.files(dir_path, pattern = "\\.nc$", full.names = TRUE)
  if (length(nc_files) == 0) {
    message("No .nc files found in the directory!")
    return(character(0))
  }
  
  broken_files <- nc_files[sapply(nc_files, is_nc_broken)]
  return(broken_files)
}
```

```{r}
broken_files <- get_broken_nc_files(sw_in_450m_tile_dir)
# Print broken file names
print(broken_files)

```

```{r}
```

```{r}
```

```{r}
```
