---
title: "Testing the Correlation Between Vegetation Height and Slope-Aspect Radiation"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(terra)       # Raster handling
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(meteoland)   # Radiation functions
library(fs)
library(multidplyr)

source(here::here("config.R"))
# source(here::here("R/helpers.R"))
source(here::here("R/batch_process_rasters.R"))
# source(here::here("R/calc_sw_in.R")) # SPLASH
source(here::here("R/compute_annual_radiation.R")) # meteoland
```

---

## Introduction

Global calculations of slope-aspect incident slolar radiation can be very large and time-consuming.
Therefore, it is useful to first run the analysis on a single tile to:

1. Verify that the code works correctly.
2. Reduce computation time for testing and debugging.

--- 

## Implementation

### Read DEM file path
```{r}

# 1️⃣ 获取所有 *_DEM.tif 文件路径
dem_30m_copernicus_dir <- "/data_2/scratch/ting/veg_topo_data/data_raw/copernicus_dem_30m/copernicus_dem_30m"
dem_files <- fs::dir_ls(
  path = dem_30m_copernicus_dir,
  glob = "*_DEM.tif",
  recurse = TRUE
)

length(dem_files)  # 总文件数，检查是否 26450
```


### Load DEM and Calculate slope & aspect 

```{r}
tictoc::tic()
# Read DEM (GeoTIFF) and target raster 
dem <- rast(dem_files[1])

# Calculate slope and aspect (in degrees)
slope <- terrain(dem, v="slope", unit="degrees")
aspect <- terrain(dem, v="aspect", unit="degrees")

# target layer for aggregation
# Crop TWI to DEM extent first, otherwise resampling DEM to TWI would expand it to global extent
twi <- rast(twi_450m_mosaic_clean_path)
res_tar <- res(twi)

# Aggregate to 450m resolution as TWI data from Marthew et. al. 
# Without resampling
aligned <- batch_process_rasters(list(dem = dem, slope = slope, aspect = aspect), res_tar = res_tar, if_resample = FALSE)
tictoc::toc()

```
---

## Extract data from raster

```{r}
df <- as.data.frame(aligned[["dem"]], xy = TRUE) |> 
  dplyr::left_join(
    as.data.frame(aligned[["slope"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  dplyr::left_join(
    as.data.frame(aligned[["aspect"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  as_tibble() |> 
  drop_na()

names(df) = c("lon", "lat", "dem", "slope" ,"aspect")
```

### calculate incident solar radiation by `meteoland::radiation_potentialRadiation`

#### no parallel

```{r}
source(here::here("R/compute_annual_radiation.R")) # meteoland
tictoc::tic()
df$meteoland_rad1 <- sapply(1:nrow(df), function(i){
  compute_annual_radiation(
    lat_deg = df$lat[i],
    slope_deg = df$slope[i],
    aspect_deg = df$aspect[i],
    year = 2020
  )
})
tictoc::toc()
```


#### parallel

```{r}
source(here::here("R/compute_annual_radiation.R")) # meteoland

tictoc::tic()
# ncores <- parallel::detectCores() - 1 
ncores <- 4
cl <- multidplyr::new_cluster(ncores) |>
      multidplyr::cluster_assign(
        compute_annual_radiation = compute_annual_radiation,
        deg2rad = function(d) d * pi / 180
      )

tmp <- df |>
  dplyr::group_by(id = row_number()) |>
  tidyr::nest(input = c(lat, slope, aspect)) |> 
  multidplyr::partition(cl) |> 
  dplyr::mutate(r_toa = purrr::map(input, 
                                   ~compute_annual_radiation(
                                     lat_deg = .x$lat[[1]],
                                     slope_deg = .x$slope[[1]], 
                                     aspect_deg = .x$aspect[[1]], 
                                     year = 2020
                                   )
  )) |>
  dplyr::collect() |>
  dplyr::ungroup() |>
  tidyr::unnest(c(r_toa, input))

tictoc::toc()
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/05_re2_cal_rad_450m_sing_tile_test.Rmd"))
```
