---
title: "Comparison of Incident Radiation Estimates from SPLASH 2.0 and meteoland"
subtitle: "Aletsch glacier"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(terra)       # Raster handling
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(lubridate)
library(meteoland)   # Radiation functions
library(multidplyr)
library(purrr)
library(furrr)

source(here::here("config.R"))
source(here::here("R/raster_preprocess_save.R"))
source(here::here("R/aggregate_topography.R"))
source(here::here("R/extent_to_tile_ids.R"))

source(here::here("R/helpers.R")) # SPLASH
source(here::here("R/calc_sw_in.R")) # SPLASH
source(here::here("R/compute_annual_radiation.R")) # meteoland

source(here::here("R/create_spatial_windows.R"))
source(here::here("R/calculate_correlation_bywin.R"))

source(here::here("R/plot_density_grid.R"))
```


# Introduction

Topography strongly influences incoming solar radiation, which in turn drives systematic variation in vegetation productivity along topographic gradients.

To test the accuracy and applicability of code adapted from SPLASH 2.0 (Sandoval et al.), we compared its estimates with those from the published R package {meteoland}.

This comparison allows us to evaluate the feasibility of SPLASH-based radiation calculations relative to an established implementation.

---


## Get slope and aspect data with 450m resolution

Load DEM 30m

```{r}
# Read DEM (GeoTIFF) and target raster 

# reduce extent to example region
# extent <- ext(-154.5,-154.25, 67.25, 67.5)
# extent <- ext(-154.5,-154, 67.25, 67.75)
# extent <- ext(-70, -67.5,-30,-29 ) # South America
# extent <- ext(6, 8, 45.5, 46.5 ) # Alp region
# extent <- ext(7.8, 7.9, 46.4, 46.5) # Aletsch glacier - beni

extent <- ext(7, 8, 46, 47) # Aletsch glacier
# Load and crop dem from tiles
dem_crop <- extent_to_tile_ids(extent, tile_size = 1,  return_raster = TRUE, source = "copernicus_dem_30m", tiles_dir = dem_30m_copernicus_dir)
names(dem_crop) = "dem"

# plot(dem_crop)
```

Load vegetation Data

```{r}

# load and crop vegh to calculate r(R~H) 
vegh <- rast(vegh_450m_mosaic_path)
vegh_crop <- terra::crop(vegh, terra::ext(dem_crop)) 

```

Load TWI data

```{r}

# target layer for aggregation
# Crop TWI to DEM extent first, otherwise resampling DEM to TWI would expand it to global extent
twi <- rast(twi_450m_mosaic_clean_path)
twi_crop <- terra::crop(twi, terra::ext(dem_crop)) # need to crop it first, otherwise the output extent of resampled one is global.
res_tar <- res(twi)
```

Calculate and aggregate slope and aspect (in degrees)

```{r}
tictoc::tic()
# Calculate slope and aspect (in degrees) and aggregate to 450m resolution as TWI data from Marthew et. al.
aligned <- aggregate_topography(dem = dem_crop, target = twi_crop, if_resample = TRUE)
tictoc::toc()
# aligned
```

---

## Extract data from raster

```{r}
df <- as.data.frame(aligned[["dem"]], xy = TRUE) |> 
  dplyr::left_join(
    as.data.frame(aligned[["slope"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  dplyr::left_join(
    as.data.frame(aligned[["aspect"]], xy = TRUE),
    by = c("x", "y")
  ) |> 
  dplyr::left_join(
    as.data.frame(vegh_crop, xy = TRUE),
    by = c("x", "y")
  ) |>
  as_tibble() |> 
  drop_na()

names(df) = c("lon", "lat", "dem", "slope" ,"aspect", "vegh")
```

---

## Input Data Visualization

```{r,fig.width=12,fig.height=12}

p_dem <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["dem"]]) +
  scale_fill_viridis_c() +
  labs(title = "Elevation (m)", fill = "Elevation") +
  theme_bw()

p_slope <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["slope"]]) +
  scale_fill_viridis_c() +
  labs(title = "Slope (°)", fill = "Slope") +
  theme_bw()

# Define a smooth circular color palette
my_colors <- c("navy", "#88ccee", "white", "#ff8888", "red", "#ff8888", "white", "#88ccee", "navy")
my_values <- seq(0, 360, length.out = length(my_colors))/360  # normalize 0-360 to 0-1

p_aspect <- ggplot() +
  tidyterra::geom_spatraster(data = aligned[["aspect"]]) +
  scale_fill_gradientn(
    colors = my_colors,
    values = my_values,
    limits = c(0, 360),
    oob = scales::squish
  ) +
  labs(title = "Aspect (°)", fill = "Aspect") +
  theme_bw()

p_vegh <- ggplot() +
  tidyterra::geom_spatraster(data = vegh_crop) +
  scale_fill_viridis_c() +
  labs(title = "Vegetation Heigh (m)", fill = "vegh") +
  theme_bw()

# print
p_dem + p_vegh + p_slope + p_aspect
```
```{r}
rm(dem_crop, twi, twi_crop, vegh, vegh_crop, aligned); gc()
```

# Comparision

---

## SPLASH 2.0

```{r SPLASH2}
source(here::here("R/calc_sw_in.R")) # SPLASH

tictoc::tic()

# Parallel Configuration
ncores <- 8
plan(multisession, workers = ncores)

# Chunk Processing
chunk_size <- 5000  # rows per chunk, adjust based on memory
chunks <- split(df, ceiling(seq_len(nrow(df)) / chunk_size))

# Parallel Calculation for Each Chunk
df <- future_map_dfr(chunks, function(chunk) {
  
  sw_in_uneven <- calc_sw_in(chunk$lat, chunk$slope, chunk$aspect, year = 2020)
  sw_in_flat <- calc_sw_in(chunk$lat, rep(0, nrow(chunk)), rep(0, nrow(chunk)), year = 2020)
  
  chunk |>
    mutate(
      sw_in_uneven = sw_in_uneven,
      sw_in_flat   = sw_in_flat,
      splash_rad = sw_in_uneven,
      index_terrain = sw_in_uneven / sw_in_flat
    )
})

# ---- End Parallel Processing ----
plan(sequential)  # restore sequential execution

tictoc::toc()
```

---

## `meteoland::radiation_potentialRadiation`

```{r meteoland}

tictoc::tic()
# ncores <- parallel::detectCores() - 1 
ncores <- 8
cl <- multidplyr::new_cluster(ncores) |>
      multidplyr::cluster_assign(
        compute_annual_radiation = compute_annual_radiation,
        deg2rad = function(d) d * pi / 180
      )

df <- df |>
  dplyr::group_by(id = row_number()) |>
  tidyr::nest(input = c(lat, slope, aspect)) |> 
  multidplyr::partition(cl) |> 
  dplyr::mutate(meteoland_rad = purrr::map(input, 
                                   ~compute_annual_radiation(
                                     lat_deg = .x$lat[[1]],
                                     slope_deg = .x$slope[[1]], 
                                     aspect_deg = .x$aspect[[1]], 
                                     year = 2020
                                   )
  )) |>
  dplyr::collect() |>
  dplyr::ungroup() |>
  tidyr::unnest(c(meteoland_rad, input))

tictoc::toc()
```


## Scatterplot Comparison

```{r,fig.width=6,fig.height=4}
colnames(df)
ggplot(df, aes(x = splash_rad, y = meteoland_rad)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    x = "SPLASH Radiation",
    y = "Meteoland Radiation",
    title = "Comparison of SPLASH vs Meteoland Radiation"
  ) +
  theme_bw()
```

## Density Distributions

```{r,fig.width=6,fig.height=4}
df_long <- df |>
  pivot_longer(cols = c(splash_rad, meteoland_rad),
               names_to = "source",
               values_to = "radiation")

ggplot(df_long, aes(x = radiation, fill = source, color = source)) +
  geom_density(alpha = 0.4) +
  labs(x = "Radiation", y = "Density", title = "Density distribution of SPLASH vs Meteoland") +
  theme_minimal()
```

## Spatial Visualization

```{r, fig.width=12,fig.height=6}
p_splash <- ggplot(df, aes(x = lon, y = lat, fill = splash_rad)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "SPLASH Radiation",
    title = "Spatial Distribution of SPLASH Radiation"
  ) +
  coord_fixed() +
  theme_bw()

p_meteoland <- ggplot(df, aes(x = lon, y = lat, fill = meteoland_rad)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "meteoland Radiation",
    title = "Spatial Distribution of meteoland Radiation"
  ) +
  coord_fixed() +
  theme_bw()

# print
p_dem + p_vegh + p_splash + p_meteoland
```

# Calculate r(R~H)

In high-latitude regions, vegetation on the south slope, which receives stronger incident solar radiation, shows lower production efficiency than on the north slope. This is contrary to the framework proposed by Fan et al. (2019).


```{r}
gc()

dwin <- 0.05  # ~500m window at equator
df_win <- create_spatial_windows(df, 
                                 coord_vars = c("lon", "lat"),
                                 value_vars = c( "dem", "slope", "aspect", "splash_rad", "meteoland_rad", "vegh"),
                                 dwin = dwin)  
colnames(df_win)

df_cor <- calculate_correlation_bywin(df_win, 
                                      x = "splash_rad",
                                      y = "vegh")            

colnames(df_cor)
```


```{r,fig.width=12,fig.height=6}
vmin <- min(df_cor$correlation, na.rm = TRUE)
vmax <- max(df_cor$correlation, na.rm = TRUE)

p_cor <- ggplot(df_cor, aes(x = lon_mid, y = lat_mid, fill = correlation)) +
  geom_tile() +
  scico::scale_fill_scico(
      palette = "vik",
      limits = c(vmin, vmax),
      midpoint = 0,
      name = expression(r[R*","*H]),
      na.value = NA
    ) +
  labs(
    x = "Longitude",
    y = "Latitude",
    fill = "r(R~H)",
    title = "r(R~H)"
  ) +
  coord_fixed() +
  theme_bw()


p_meteoland + p_cor
```
```{r}
# colnames(df)
cols <- c( "vegh", "splash_rad",  "dem","slope", "aspect", "meteoland_rad", "lon", "lat")
combined_plot <- plot_density_grid(df, cols, nrow = 3, save_path = NULL)
combined_plot
```

```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/05_re1_compare_radiation_methods.Rmd"))
```
