---
title: "r_H_R Calculation on defined regions"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# ------Load required libraries-------------------------------------------------

library(terra)     # For handling raster data
library(furrr)
library(dplyr)
library(ggmap)
library(tidyterra)
library(doParallel)    # inner parallelism
library(foreach)

library(khroma)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
library(rnaturalearth)
library(sf)

# ------Load configuration and helper functions---------------------------------

source(here::here("config.R"))
source(here::here("R/create_spatial_windows.R"))
source(here::here("R/calculate_correlation_bywin.R"))
source(here::here("R/mosaic_tiles.R"))
source(here::here("R/plot_dem.R"))
source(here::here("R/plot_aspect.R"))
source(here::here("R/plot_slope.R"))
source(here::here("R/plot_vegh.R"))
source(here::here("R/plot_r_H_R.R"))
source(here::here("R/plot_sw_in.R"))
source(here::here("R/plot_terrain_effect.R"))
source(here::here("R/plot_hex_scatter.R"))
source(here::here("R/plot_single_sample_location.R"))
source(here::here("R/plot_google_img.R"))
# source(here::here("R/plot_cci_land_cover.R"))
source(here::here("R/plot_fused.R"))

source(here::here("R/raster_preprocess_save.R"))
source(here::here("R/aggregate_topography.R"))
source(here::here("R/extent_to_tile_ids.R"))

source(here::here("R/helpers.R")) # SPLASH
source(here::here("R/calc_sw_in.R")) # SPLASH
```


---

## Task

To verify the correctness of the split-window correlation analysis, all input and intermediate datasets should be visualized to ensure the accuracy of the workflow.

We manually selected several well-known regions, including those mentioned by Fan et al. (2019), for this validation.
The evaluation will be conducted at two spatial scales:

1. Coarse resolution (5 km): Pearson’s r will be computed using the radiation index and vegetation height aggregated to 450 m.
2. Fine resolution (500 m): Pearson’s r will be calculated using the radiation index and vegetation height data at 30 m resolution.

This setup allows us to assess whether the analysis performs consistently across different spatial resolutions and datasets.

---

## Define regions

```{r}

tiles_info <- data.frame(
  tile_id = c(
    "Aletsch_glacier",
    "3a_arid_western_Texas",
    "3b_seasonal_arid_California",
    "3c_seasonal_arid_Idaho",
    "3d_winter_snowpacks_Mount_Cramer",
    "3e_high_latitude_southwest_Yukon",
    "3f_boreal_forest_Alaska"
  ),
  ymin = c(46, 31.5, 37.5, 43, 43.8, 61.5, 64),
  ymax = c(47, 32.5, 38.5, 44, 44.1, 62.5, 65),
  xmin = c(7, -105, -120.5, -115.8, -115.5, -138.5, -148),
  xmax = c(8, -104.5, -119.5, -114.8, -114.5, -137.5, -147)
)

# center location
tiles_info$xmid <- (tiles_info$xmax + tiles_info$xmin) / 2
tiles_info$ymid <- (tiles_info$ymax + tiles_info$ymin) / 2
```

---

## Global analysis on validation regions

```{r, eval=FALSE}
process_tile_win_5km <- function(tile_row,
                         # sw_in_terrain_effect_450m_path = sw_in_terrain_effect_450m_path,
                         # vegh_450m_mosaic_path = vegh_450m_mosaic_path,
                         # dem_30m_copernicus_dir = dem_30m_copernicus_dir,
                         # r_H_R_5km_path = r_H_R_5km_path,
                         out_dir = "data/figures",
                         text_size = 12,
                         fig_width = 14,
                         fig_height = 20
) {
  
  # tile info
  tile_id <- tile_row$tile_id
  tile_extent <- terra::ext(tile_row$xmin,
                            tile_row$xmax,
                            tile_row$ymin,
                            tile_row$ymax)
  tile_xmid <- (tile_row$xmin + tile_row$xmax) / 2
  tile_ymid <- (tile_row$ymin + tile_row$ymax) / 2

  tictoc::tic(paste0("Processing tile: ", tile_id))
  
  # ---- Load rasters ----
  terrain_effect_r <- terra::rast(sw_in_terrain_effect_450m_path)
  terrain_effect_rc <- terra::crop(terrain_effect_r, tile_extent)
  
  vegh_r <- terra::rast(vegh_450m_mosaic_path)
  vegh_rc <- terra::crop(vegh_r, tile_extent)
  
  # Stack and name layers
  stacked <- c(terrain_effect_rc, vegh_rc)
  names(stacked) <- c("rin", "vegh")
  
  # Load DEM
  dem_crop <- extent_to_tile_ids(
    tile_extent,
    tile_size = 1,
    return_raster = TRUE,
    source = "copernicus_dem_30m",
    tiles_dir = dem_30m_copernicus_dir
  )
  names(dem_crop) <- "dem"
  
  # ---- Aggregate DEM and calculate slope/aspect ----
  aligned <- aggregate_topography(dem_crop,
                                  res_tar = NULL,
                                  target = vegh_rc,
                                  if_resample = TRUE)
  
  # ---- Spatial windows and correlation ----
  df_win <- create_spatial_windows(stacked, value_vars = c("rin", "vegh"))
  df_cor <- calculate_correlation_bywin(df_win, x = "rin", y = "vegh")
  cor_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "correlation")], type = "xyz", crs = "EPSG:4326")
  
  # ---- Generate plots ----
  p_dem <- plot_dem(aligned[["dem"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_slope <- plot_slope(aligned[["slope"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_aspect <- plot_aspect(aligned[["aspect"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_vegh <- plot_vegh(vegh_rc, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_rin <- plot_terrain_effect(terrain_effect_rc, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_r_H_R <- plot_r_H_R(cor_r, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_r_H_R2 <- plot_r_H_R(r_H_R_5km_path, extent = tile_extent, title_text = "Pre. Pearson's r (H ~ R)", text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_google <- plot_google_img(extent = tile_extent)
  p_fused <- plot_fused(fused_5km_file, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)

  p_scatter <- plot_hex_scatter(df_win, x_var = "rin", y_var = "vegh",
                                x_text = "Radiation index", y_text = "Vegetation height (m)",
                                text_size = text_size)
  p_location <- plot_single_sample_location(tile_xmid, tile_ymid, tile_id, text_size = text_size)
  # ---- Combine plots ----
  final_plot <- ((p_dem + p_slope + p_aspect) /
                 (p_r_H_R + p_vegh + p_rin) /
                 (p_r_H_R2 + p_location + p_scatter))  /
                 (p_google + p_fused) +
                 plot_annotation(title = tile_id) +
                 plot_layout(heights = c(1, 1, 1, 1))
  
  # ---- Save plot ----
  out_file <- here::here(file.path(out_dir, paste0("05_", tile_id, "_win_5km_plots.png")))
  ggsave(filename = out_file, plot = final_plot, width = fig_width, height = fig_height, dpi = 600)
  
  # ---- Memory cleanup ----
  rm(terrain_effect_r, terrain_effect_rc, vegh_r, vegh_rc, stacked,
     dem_crop, aligned, df_win, df_cor, cor_r,
     p_dem, p_slope, p_aspect, p_vegh, p_rin, p_fused, p_google, p_r_H_R, p_r_H_R2, p_scatter, p_location, final_plot)
  gc(verbose = FALSE)
  
  tictoc::toc()
  return(out_file)
}


for (i in seq_len(nrow(tiles_info))) {
  process_tile_win_5km(tiles_info[i, ])
}
```

---

## Regional Analysis on validation regions

```{r, eval=FALSE}
process_tile_win_500m <- function(tile_row,
                         # sw_in_terrain_effect_450m_path = sw_in_terrain_effect_450m_path,
                         # vegh_450m_mosaic_path = vegh_450m_mosaic_path,
                         # dem_30m_copernicus_dir = dem_30m_copernicus_dir,
                         # r_H_R_5km_path = r_H_R_5km_path,
                         out_dir = "data/figures",
                         text_size = 12,
                         fig_width = 14,
                         fig_height = 20
) {
  
    # tile info
  tile_id <- tile_row$tile_id
  tile_extent <- terra::ext(tile_row$xmin,
                            tile_row$xmax,
                            tile_row$ymin,
                            tile_row$ymax)
  tile_xmid <- (tile_row$xmin + tile_row$xmax) / 2
  tile_ymid <- (tile_row$ymin + tile_row$ymax) / 2

  tictoc::tic(paste0("Processing tile: ", tile_id))
  
   # Load TWI Raster ---
    twi_r <- terra::rast(twi_30m_path)
    twi_rc <- crop(twi_r, tile_extent, snap = "out")
    names(twi_rc) <- "twi"
    
    twi_rc

    rm(twi_r);gc()

    # Load and Prepare Vegetation Height Raster ---
    vegh_rc <- extent_to_tile_ids(tile_extent, tile_size = 3, return_raster = TRUE,
                                  source = "lang_vegh_10m", tiles_dir = vegh_10m_tiles_dir)
    names(vegh_rc) <- "vegh"

    # Set 0 as NA value (0m canopy height represents not vegetated or water according to Lang et al. (2019))
    # Aggregates using TWI data from Ho et al. (2015)
    # Return the saved path
    vegh_rr <- raster_preprocess_save(
      input = vegh_rc,
      target = twi_rc,
      na_value = 0,
      fun = mean,
      varname = "vegh",
      if_aggregate = TRUE,
      if_resample = TRUE,
      if_mask = TRUE,
      if_return_raster = TRUE
    )

  # Load DEM
  dem_crop <- extent_to_tile_ids(
    tile_extent,
    tile_size = 1,
    return_raster = TRUE,
    source = "copernicus_dem_30m",
    tiles_dir = dem_30m_copernicus_dir
  )
  names(dem_crop) <- "dem"
  
  # ---- Aggregate DEM and calculate slope/aspect ----
  aligned <- aggregate_topography(dem_crop,
                                  res_tar = NULL,
                                  target = twi_rc,
                                  if_resample = TRUE)
  
  # Extract + join
    df <- as.data.frame(aligned[["dem"]], xy = TRUE) |>
      left_join(as.data.frame(aligned[["slope"]], xy = TRUE), by = c("x", "y")) |>
      left_join(as.data.frame(aligned[["aspect"]], xy = TRUE), by = c("x", "y")) |>
      tibble::as_tibble() |>
      drop_na()
    names(df) <- c("lon", "lat", "dem", "slope", "aspect")

   if (nrow(df) == 0) {
      warning(sprintf("No valid cells after drop_na for %s", file))
      return(FALSE)
    }
    
    
    # clean old cluster
    try({
      if (exists("cl")) stopCluster(cl)
      closeAllConnections()
    }, silent = TRUE)
    
    # Inner parallelism
    num_cores <- 8
    cl <- makeCluster(num_cores)
    registerDoParallel(cl)

    # Chunk Processing
    chunk_size <- 5000  # rows per chunk, adjust based on memory
    chunks <- split(df, ceiling(seq_len(nrow(df)) / chunk_size))

    # Parallel Calculation for Each Chunk - Direct assignment
    df_calc <- foreach(
      chunk = chunks,
      .combine = bind_rows,
      .packages = c("dplyr"),
      .export   = c("calc_sw_in_daily", "calc_sw_in", "julian_day",
                    "berger_tls", "dcos", "dsin")
    ) %dopar% {

      # Calculate sw_in_uneven and sw_in_flat for entire chunk
      sw_in_uneven <- calc_sw_in(chunk$lat, chunk$slope, chunk$aspect, year = 2020)
      sw_in_flat <- calc_sw_in(chunk$lat, rep(0, nrow(chunk)), rep(0, nrow(chunk)), year = 2020)

      # Combine results back to dataframe
      chunk |>
        mutate(sw_in_uneven = sw_in_uneven,
         sw_in_flat = sw_in_flat,
         rin = sw_in_uneven / sw_in_flat)

    }

    stopCluster(cl)
    registerDoSEQ()
    
    # Build rasters
    crs_out <- terra::crs(aligned[["dem"]])
    sw_in_uneven <- terra::rast(df_calc[, c("lon", "lat", "sw_in_uneven")], type = "xyz", crs = crs_out)
    sw_in_flat <- terra::rast(df_calc[, c("lon", "lat", "sw_in_flat")], type = "xyz", crs = crs_out)
    rin <- terra::rast(df_calc[, c("lon", "lat", "rin")], type = "xyz", crs = crs_out)
    rin <- terra::resample(rin, twi_rc, method = "near")
  # --- Stack Rasters ---
    stacked <- c(rin, vegh_rr)
    names(stacked) <- c("rin", "vegh")
    
  # ---- Spatial windows and correlation ----
  df_win <- create_spatial_windows(stacked, value_vars = c("rin", "vegh"), dwin = 0.005)
  df_cor <- calculate_correlation_bywin(df_win, x = "rin", y = "vegh")
  cor_r <- terra::rast(df_cor[, c("lon_mid", "lat_mid", "correlation")], type = "xyz", crs = "EPSG:4326")
  
  # ---- Generate plots ----
  p_dem <- plot_dem(aligned[["dem"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_slope <- plot_slope(aligned[["slope"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_aspect <- plot_aspect(aligned[["aspect"]], extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_vegh <- plot_vegh(vegh_rr, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_rin <- plot_terrain_effect(rin, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_r_H_R <- plot_r_H_R(cor_r, extent = tile_extent,  title_text = "500m: Pearson's r (H ~ R)", text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_r_H_R2 <- plot_r_H_R(r_H_R_5km_path, extent = tile_extent, title_text = "5km: Pearson's r (H ~ R)", text_size = text_size, x_step = 0.5, y_step = 0.5)
  p_google <- plot_google_img(extent = tile_extent)
  p_fused <- plot_fused(fused_5km_file, extent = tile_extent, text_size = text_size, x_step = 0.5, y_step = 0.5)

  p_scatter <- plot_hex_scatter(df_win, x_var = "rin", y_var = "vegh",
                                x_text = "Radiation index", y_text = "Vegetation height (m)",
                                text_size = text_size)
  p_location <- plot_single_sample_location(tile_xmid, tile_ymid, tile_id, text_size = text_size)
  # ---- Combine plots ----
  final_plot <- ((p_dem + p_slope + p_aspect) /
                 (p_r_H_R + p_vegh + p_rin) /
                 (p_r_H_R2 + p_location + p_scatter))  /
                 (p_google + p_fused) +
                 plot_annotation(title = tile_id) +
                 plot_layout(heights = c(1, 1, 1, 1))
  
  # ---- Save plot ----
  out_file <- here::here(file.path(out_dir, paste0("05_", tile_id, "_win_500m_plots.png")))
  ggsave(filename = out_file, plot = final_plot, width = fig_width, height = fig_height, dpi = 600)
  
  # ---- Memory cleanup ----
  # list all objects need to be remove
  cleanup_vars <- c(
   "twi_r", "twi_rc",
   "vegh_rc", "vegh_rr",
   "dem_crop", "aligned",
    "df", "chunks", "df_calc",
   "stacked_r", "df_win", "df_cor", "cor_r",
    "p_dem", "p_slope", "p_aspect", "p_vegh", "p_rin",
    "p_r_H_R", "p_r_H_R2", "p_google", "p_fused",
   "p_scatter", "p_location", "final_plot"
  )

  # remove existed variables
  cleanup_vars <- cleanup_vars[cleanup_vars %in% ls()]

  rm(list = cleanup_vars)
  gc(verbose = FALSE)
  
  tictoc::toc()
  return(out_file)
}

for (i in seq_len(nrow(tiles_info))) {
  process_tile_win_500m(tiles_info[i, ])
}
```


```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/05_re3.Rmd"))
```
