---
title: "Pearson's r (H~R) Calculation in sample regions"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

## Task

The Pearson correlation analysis between vegetation height and topographic wetness index for all automaticlly selected regions has been completed in `~/veg_topo/analysis/06_1_regB_r_H_R.R`.

This markdown file focuses on visualizing the plots from the script.

```{r}
library(terra)
# Automatically select configuration file
hostname <- trimws(tolower(system("hostname", intern = TRUE)))
if (hostname == "dash") {
  message("💻 Detected Worksation: dash → using config.R")
  source(here::here("config.R"))
} else {
  message("🖥️ Detected HPC environment (", hostname, ") → using config_ubelix.R")
  source(here::here("config_ubelix.R"))
}
```

### Final Sample

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_.*_1_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### low_latitude_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_low_latitude_flat_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### low_latitude_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_low_latitude_rugged_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### relatively_low_latitude_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_relatively_low_latitude_flat_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### relatively_low_latitude_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_relatively_low_latitude_rugged_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### mid_latitude_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_mid_latitude_flat_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### mid_latitude_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_mid_latitude_rugged_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### relatively_high_latitude_flat_relief

```{r}

# list all figures

files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_relatively_high_latitude_flat_relief.*_plots\\.png$", full.names = TRUE)

# Print

cat("Found", length(files), "figures:\\n")

# print

knitr::include_graphics(files)
```

### relatively_high_latitude_rugged_relief

```{r}
# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_relatively_high_latitude_rugged_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### high_latitude_flat_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_high_latitude_flat_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

# print
knitr::include_graphics(files)
```

### high_latitude_rugged_relief

```{r}

# list all figures
files <- list.files(path = here::here("data/figures"), pattern = "^06_regB_high_latitude_rugged_relief.*_plots\\.png$", full.names = TRUE)

# Print 
cat("Found", length(files), "figures:\n")

knitr::include_graphics(files)
```

```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/06_re1.Rmd"))
```
