---
title: "Checking the complement and numbers of calculated incident solar radiation"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

## Task

We perform a parallelized calculation of **surface solar radiation** over **26,450 1×1 degree tiles**. Some outputs may be incomplete or corrupted due to errors during processing.

Goals:

1. **Verify outputs** by checking the number of generated files.
2. **Assess quality** to detect incomplete or broken results.
3. **Identify missing or broken tiles** and print their **tile IDs** for follow-up.

---

```{r setup, include=FALSE}
library(terra)  # For raster operations
library(fs)     # File system utilities
library(ncdf4)  # For NetCDF file handling
source(here::here("R/config.R"))  # Optional config file
```

---

## Input Data Information

```{r}
# List all DEM input files recursively
dem_files_all <- fs::dir_ls(
  path = dem_30m_copernicus_dir,
  glob = "*_DEM.tif",
  recurse = TRUE
)

message(sprintf("Found %d DEM tiles", length(dem_files_all)))
```

---

## Output Data Information and Missing Files

```{r}
# List processed SW_IN NetCDF files
sw_in_uneven_files <- list.files(
  path = sw_in_uneven_450m_tile_dir,
  pattern = "*_to_sw_in_uneven_450m\\.nc$",
  recursive = TRUE,
  full.names = TRUE
)

sw_in_flat_files <- list.files(
  path = sw_in_flat_450m_tile_dir,
  pattern = "*_to_sw_in_flat_450m\\.nc$",
  recursive = TRUE,
  full.names = TRUE
)

# Extract base names to compare with DEM tiles
dem_base_names <- gsub("\\.tif$", "", basename(dem_files_all))
sw_in_uneven_base_names <- gsub("_to_sw_in_uneven_450m\\.nc$", "", basename(sw_in_uneven_files))
sw_in_flat_base_names <- gsub("_to_sw_in_flat_450m\\.nc$", "", basename(sw_in_flat_files))

# Identify missing output files
missing_sw_in_uneven <- setdiff(dem_base_names, sw_in_uneven_base_names)
missing_sw_in_flat <- setdiff(dem_base_names, sw_in_flat_base_names)

# identify missing output files indices
missing_sw_in_uneven_idx <- which(dem_base_names %in% missing_sw_in_uneven)
missing_sw_in_flat_idx <- which(dem_base_names %in% missing_sw_in_flat)

# Print summary messages in a clear format
message(sprintf("[%s] Total missing uneven SW_IN files: %d", Sys.time(), length(missing_sw_in_uneven)))
message(sprintf("[%s] Total missing flat SW_IN files: %d", Sys.time(), length(missing_sw_in_flat)))
```

```{r}
message("Indices of missing uneven SW_IN tiles:")
message(paste(missing_sw_in_uneven_idx, collapse = ", "))

message("Indices of missing flat SW_IN tiles:")
message(paste(missing_sw_in_flat_idx, collapse = ", "))
```

---

## Manual Tile Check (1:26450)

```{r}
# Function to check a single tile's DEM and SW_IN outputs
check_tile <- function(tile_index) {
  if (tile_index > length(dem_files_all)) {
    message("Tile index out of range")
    return(NULL)
  }

  dem_file <- dem_files_all[tile_index]
  dem_base_name <- gsub("\\.tif$", "", basename(dem_file))

  sw_in_uneven <- file.path(sw_in_uneven_450m_tile_dir, paste0(dem_base_name, "_to_sw_in_uneven_450m.nc"))
  sw_in_flat <- file.path(sw_in_flat_450m_tile_dir, paste0(dem_base_name, "_to_sw_in_flat_450m.nc"))

  # Check if output files exist
  if (!file.exists(sw_in_uneven) || !file.exists(sw_in_flat)) {
    message(sprintf("Output files missing for: %s", dem_base_name))
    return(NULL)
  }

  return(list(dem = dem_file, uneven = sw_in_uneven, flat = sw_in_flat))
}
```

```{r}

gc()

# Example: Visualize tile 199
tile_data <- check_tile(199)
if (!is.null(tile_data)) {
  tile_data$dem
  dem <- rast(tile_data$dem)
  dem_plot <- aggregate(dem, fact = 10) 
  uneven <- rast(tile_data$uneven)
  flat <- rast(tile_data$flat)
    
  par(mfrow = c(1, 3))
  plot(dem_plot, main = "DEM")
  plot(uneven, main = "Uneven SW_IN")
  plot(flat, main = "Flat SW_IN")
}
```

---

