---
title: "Validate vegetation height data"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(terra)
library(dplyr) # need in calculate_correlation_bywin()
library(sf) # need in get_lonlat_extent()
library(tidyr) # drop_na()
library(progress)
library(ggplot2)
library(patchwork)

source(here::here("R/config.R"))
source(here::here("R/get_lonlat_extent.R"))  # need in create_aligned_template()
source(here::here("R/create_aligned_template.R"))
source(here::here("R/create_spatial_windows.R"))
source(here::here("R/calculate_correlation_bywin.R"))
source(here::here("R/raster_preprocess_save.R"))
source(here::here("R/aggregate_topography.R"))
source(here::here("R/extent_to_tile_ids.R"))
source(here::here("R/helpers.R"))        # SPLASH
source(here::here("R/calc_sw_in.R"))     # SPLASH

source(here::here("R/plot_dem.R"))
source(here::here("R/plot_vegh.R"))
source(here::here("R/plot_twi.R"))
source(here::here("R/plot_rin.R"))

source(here::here("R/plot_google_img.R"))
source(here::here("R/plot_cor_twi_vegh.R"))
source(here::here("R/plot_cor_pval.R"))
source(here::here("R/plot_hex_scatter.R"))
source(here::here("R/plot_r_H_R.R"))

```

## Task

Due to the unignorable uncertainty in the canopy height estimates from the Lang et al. model, we aim to assess how this uncertainty affects the correlation between canopy height (H) and topographic indices (TWI and Rin). To isolate the influence of Lang et al.'s model uncertainty, we use airborne LiDAR-derived canopy height as a ground truth reference. By comparing the correlations obtained using LiDAR-derived H versus Lang et al.'s H, we can evaluate how the uncertainty in Lang et al.'s model alters the correlation with topographic indices.

---
## Input data pre

### H and TWI
```{r data_H_TWI}
# Load Lidar H data and get the value 0-70
file <- file.path(lidar_asl_dir, "32TMS.tif")
lidar <- rast(file)
lidar <- clamp(lidar, lower = 0, upper = 70, values = FALSE)

# set up align template according to the extent from Lidar H and resolution from Ho's twi (30m)
align_r <- create_aligned_template(lidar, res_out = 0.00025)
align_r

# project the Lidar H to the align template
vegh_lidar <- project(lidar, align_r, method = "average")
vegh_lidar

# load twi, crop and mask for better comparison
reg_extent <- ext(align_r)
twi_rc <- terra::mask(
  terra::crop(terra::rast(twi_30m_path), reg_extent) / 100,
  vegh_lidar
)
twi_rc

# Load lang's H, set NA value, resample and mask it
# Set 0 as NA value (0m canopy height represents not vegetated or water according to Lang et al. (2019))
# Aggregate and resample using TWI data from Ho et al. (2025)
vegh_lang <- extent_to_tile_ids(reg_extent, tile_size = 3, return_raster = TRUE,
                              source = "lang_vegh_10m", tiles_dir = vegh_10m_tiles_dir)
vegh_lang <- raster_preprocess_save(
  input = vegh_lang,
  target = vegh_lidar,
  na_value = 0,
  fun = mean,
  varname = "vegh",
  if_aggregate = TRUE,
  if_resample = TRUE,
  if_mask = TRUE,
  if_return_raster = TRUE
)
vegh_lang
```

### DEM and Rin
```{r data_dem_srad}
# --- Elevation and radiation Raster ---
dem_rc <- extent_to_tile_ids(reg_extent, tile_size = 1, return_raster = TRUE,
                             source = "copernicus_dem_30m", tiles_dir = dem_30m_copernicus_dir)

# Get slope and aspect and Aggregate to ~450m
aligned <- aggregate_topography(
  dem_rc,
  res_tar = NULL,
  target = align_r,
  if_aggregate = FALSE,
  if_resample = TRUE
)

# Extract + join
df_topo <- as.data.frame(aligned[["dem"]], xy = TRUE) |>
      left_join(as.data.frame(aligned[["slope"]], xy = TRUE), by = c("x", "y")) |>
      left_join(as.data.frame(aligned[["aspect"]], xy = TRUE), by = c("x", "y")) |>
      tibble::as_tibble() |>
      drop_na()
names(df_topo) <- c("lon", "lat", "dem", "slope", "aspect")

if (nrow(df_topo) == 0) {
  warning(sprintf("No valid cells after drop_na for %s", file))
}

# Workstation2
tictoc::tic() 
batch_size <- 100000
n_rows <- nrow(df_topo)

# 1. 计算总批次数
n_batches <- ceiling(n_rows / batch_size)

sw_in_uneven <- numeric(n_rows)
sw_in_flat <- numeric(n_rows)

# 2. 创建 R 自带进度条
pb <- txtProgressBar(min = 0, max = n_batches, style = 3)

for(i in seq(1, n_rows, by = batch_size)) {
  idx <- i:min(i + batch_size - 1, n_rows)
  current_batch <- ceiling(i / batch_size)
  
  sw_in_uneven[idx] <- calc_sw_in(df_topo$lat[idx], 
                                   df_topo$slope[idx], 
                                   df_topo$aspect[idx], 
                                   year = 2020)
  
  sw_in_flat[idx] <- calc_sw_in(df_topo$lat[idx], 
                                 rep(0, length(idx)), 
                                 rep(0, length(idx)), 
                                 year = 2020)
  
  # 3. 更新进度条
  setTxtProgressBar(pb, current_batch)
}

close(pb) # 关闭进度条
tictoc::toc()

# # UBELIX, need enough memory >500
# # 计算辐射（向量化操作，处理所有行）
# sw_in_uneven <- calc_sw_in(df$lat, df$slope, df$aspect, year = 2020)
# sw_in_flat <- calc_sw_in(df$lat, rep(0, nrow(df)), rep(0, nrow(df)), year = 2020)
# # 


# calculate rin
df_calc <- df_topo |>
  mutate(rin = sw_in_uneven / sw_in_flat)

# Build rin rasters
rin <- terra::rast(df_calc[, c("lon", "lat", "rin")], type = "xyz", crs = crs(align_r))
rin <- terra::extend(rin, align_r) |> terra::mask(vegh_lidar)

dem <- aligned[["dem"]] |> terra::mask(vegh_lidar)
```

### Plot input data
#### layout pre
```{r retheme}
text_size = 14
x_step = 0.5
y_step = 0.5

# --- reset theme for plots, change the layout for better visualization ---
re_theme0 <- list(
  ggplot2::theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    panel.background = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA),
    legend.box.background = element_rect(fill = NA, color = NA),
    
    legend.text = ggplot2::element_text(size = text_size * 0.9,
                                        angle = 90,
                                        hjust = 0.5, vjust = 0.5,
                                        margin = margin(r = 0, b = 0, l = 0)),
    legend.title = ggplot2::element_text(size = text_size,
                                         angle = 0,   hjust = 0, vjust = 1 ),
    legend.margin = margin(0, 0, 0, 0),
    legend.box.margin = margin(-45, 0, 0, 0),
    axis.text.x = ggplot2::element_text(
      angle = 90,
      size = text_size * 0.8,
      hjust = 0.5,
      vjust = 0.5,
      margin = margin(t = 0, b = 0),
    ),
    axis.text.y = ggplot2::element_text(
      angle = 90,
      size = text_size * 0.8,
      hjust = 0.5,
      vjust = 0.5,
      margin = margin(r = 2, l = 0)
    ),
    panel.spacing = unit(0, "cm"),
    panel.border = ggplot2::element_rect(linewidth = 0.3, fill = NA),
    plot.title = ggplot2::element_text(
      size = text_size * 1.2,
      face = "plain",
      margin = margin(b = 0)
    ),
    plot.title.position = "panel"
  )
)

re_theme <- list(
  guides(fill = guide_colorbar(barwidth = 8, barheight = 0.8)),
  ggplot2::theme(  axis.title.x = ggplot2::element_blank(),
                   axis.title.y = ggplot2::element_blank()),
  re_theme0
)
```

#### plotting
```{r plot_input}

p_vegh_lang <- plot_vegh(vegh_lang, extent = reg_extent, title_text = expression("   30-m " * italic(H)), text_size = text_size, x_step = x_step, y_step = y_step) +
  labs(tag = "d)") +
  theme(
    plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
    plot.tag.position = c(0.06, 1)
  ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 0, r = 0, b = 4, l = 0))

p_vegh_lidar <- plot_vegh(vegh_lidar, extent = reg_extent, title_text = expression("   30-m " * italic(H)), text_size = text_size, x_step = x_step, y_step = y_step) +
  labs(tag = "d)") +
  theme(
    plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
    plot.tag.position = c(0.06, 1)
  ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 0, r = 0, b = 4, l = 0))

p_dem <- plot_dem(dem, extent = reg_extent, title_text = "   30-m Elevation", text_size = text_size, x_step = x_step, y_step = y_step) +
  labs(tag = "c)") +
  theme(
    plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
    plot.tag.position = c(0.12, 1)
  ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 0, r = 20, b = 4, l = 0))

p_twi <- plot_twi(twi_rc, extent = reg_extent, title_text = "   30-m TWI", text_size = text_size, x_step = x_step, y_step = y_step) +
  labs(tag = "e)") +
  theme(
    plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
    plot.tag.position = c(0.13, 1)
  ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 4, r = 20, b = 4, l = 0))

# p_rin <- plot_rin(rin, extent = reg_extent, title_text = "   30-m Rᵢₙ",  text_size = text_size, x_step = x_step, y_step = y_step) +
#   labs(tag = "g)") +
#   theme(
#     plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
#     plot.tag.position = c(0.12, 1)
#   ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 4, r = 20, b = 4, l = 0))

final_plot <- patchwork::wrap_plots(
      p_vegh_lang, p_vegh_lidar, p_dem, patchwork::plot_spacer(),
      # p_twi, p_rin, patchwork::plot_spacer(),
      p_twi, patchwork::plot_spacer(), patchwork::plot_spacer(),

      ncol = 3, nrow = 2
    ) &
      theme(
        panel.background = element_rect(fill = "white", color = NA),
        plot.background  = element_blank(),
        legend.background = element_blank(),
        legend.box.background = element_blank(),
      )

out_file2 <- here::here(file.path(paste0("data/figures/6_01_validate_", basename(file), "_input.png")))
ggsave(filename = out_file2, plot = final_plot, width = 14, height = 8, dpi = 600)

```

---
## Corelation (H, TWI)
### Compute
```{r r_H_TWI}
stacked_lang <- c(twi_rc, rin, vegh_lang, vegh_lidar)
df_win_lang <- create_spatial_windows(stacked_lang, value_vars = c("twi", "rin","vegh_lang", "vegh_lidar"), dwin = 0.005)
df_cor_lang <- calculate_correlation_bywin(df_win_lang, x = "twi", y = "vegh_lang")
df_cor_lidar <- calculate_correlation_bywin(df_win_lidar, x = "twi", y = "vegh_lidar")

df_combined <- merge(df_cor_lang, df_cor_lidar,
                     by = c("lon_mid", "lat_mid"),
                     suffixes = c("_lang", "_lidar"))
colnames(df_combined)
```
### summary
#### text
```{r r_H_TWI_summary}
summary_cor <- data.frame(
  Variable = c("correlation_lang", "correlation_lidar"),
  Mean = c(mean(df_combined$correlation_lang, na.rm = TRUE),
           mean(df_combined$correlation_lidar, na.rm = TRUE)),
  SD = c(sd(df_combined$correlation_lang, na.rm = TRUE),
         sd(df_combined$correlation_lidar, na.rm = TRUE)),
  Min = c(min(df_combined$correlation_lang, na.rm = TRUE),
          min(df_combined$correlation_lidar, na.rm = TRUE)),
  Max = c(max(df_combined$correlation_lang, na.rm = TRUE),
          max(df_combined$correlation_lidar, na.rm = TRUE))
)
print(summary_cor)

df_combined$cor_diff <- df_combined$correlation_lang - df_combined$correlation_lidar
df_combined$abs_diff <- abs(df_combined$cor_diff)

# 两个相关性的相关性
cor_between <- cor(df_combined$correlation_lang,
                   df_combined$correlation_lidar,
                   use = "complete.obs")
cat("Corelattion between r(H_lang, TWI) and r(H_liadar, TWI):", cor_between, "\n")
```

#### figures
```{r r_H_TWI_comparison}

# 差异箱线图
boxplot(df_combined$correlation_lang, df_combined$correlation_lidar,
        names = c("Lang", "Lidar"),
        main = "Comparison of Correlations",
        ylab = "Correlation",
        col = c("lightgreen", "lightcoral"))

p1 <- ggplot(df_combined, aes(x = correlation_lang, y = correlation_lidar)) +
  geom_hex(bins = 30) +  # bins 控制六边形数量
  scale_fill_viridis_c(name = "Count", trans = "log") +  # 使用对数变换更好显示密度
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue", size = 1) +
  labs(title = "Correlation Comparison",
       x = "Lang Correlation",
       y = "Lidar Correlation") +
  theme_minimal() +
  theme(legend.position = "right")

p2 <- ggplot(df_combined, aes(x = cor_diff)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Difference Distribution", x = "Difference (Lang - Lidar)") +
  theme_minimal()

p3 <- ggplot(df_combined, aes(y = correlation_lang, x = "Lang")) +
  geom_boxplot(fill = "lightgreen") +
  geom_boxplot(aes(y = correlation_lidar, x = "Lidar"), fill = "lightcoral") +
  labs(title = "Boxplot Comparison", y = "Correlation") +
  theme_minimal()

# 合并图形

(p1) / (p2 | p3) + plot_layout(heights = c(2, 1))  +
  plot_annotation(title = "Figure: Correlation Analysis",
                  theme = theme(plot.title = element_text(hjust = 0.5)))
```


```{r r_H_TWI_plot}
p_rA_lang <- plot_cor_twi_vegh(cor_lang, extent = reg_extent,  title_text <- bquote("   500-m Pearson's " * r[.("H")*","*.("TWI")]), text_size = text_size, x_step = x_step, y_step = y_step) +
  labs(tag = "f)") +
  theme(
    plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
    plot.tag.position = c(0.06, 1)
  )
p_rA_lang
```

---
## Corelation (H, Rin)
### Compute
```{r r_H_Rin}

df_win_lang <- create_spatial_windows(stacked_lang, value_vars = c("twi", "vegh_lang", "vegh_lidar"), dwin = 0.005)
df_cor_lang <- calculate_correlation_bywin(df_win_lang, x = "twi", y = "vegh_lang")
df_cor_lidar <- calculate_correlation_bywin(df_win_lidar, x = "twi", y = "vegh")

df_combined <- merge(df_cor_lang, df_cor_lidar,
                     by = c("lon_mid", "lat_mid"),
                     suffixes = c("_lang", "_lidar"))
colnames(df_combined)
```


```{r plot_r_H_Rin}
    p_rB <- plot_r_H_R(file.path(output_dir, paste0("reg_", reg_id, "_r_H_R_30m_500m_map.nc")), extent = reg_extent, title_text = bquote("   500-m Pearson's " * r[.("H")*","*.("Rᵢₙ")]), text_size = text_size, x_step = x_step, y_step = y_step) +
      labs(tag = "h)") +
      theme(
        plot.tag = element_text(size = 14, face = "bold", hjust = 0, vjust = 1),
        plot.tag.position = c(0.06, 1)
      ) + re_theme +  ggplot2::theme(plot.margin = margin(t = 4, r = 0, b = 4, l = 0))
```
