---
title: "Model Data Preprocessing - Tile Test"
author: "Ting Tan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(terra) 
library(dplyr)
library(tidyr)
library(ggplot2)
library(arrow)
# Load custom functions
source(here::here("R/config.R"))  
source(here::here("R/create_aligned_template.R"))
source(here::here("R/mosaic_tiles.R"))
source(here::here("R/extent_to_tile_ids.R"))
source(here::here("R/raster_preprocess_save.R"))
source(here::here("R/aggregate_topography.R"))
source(here::here("R/create_spatial_windows.R"))
source(here::here("R/make_lon_label.R")) 
source(here::here("R/helpers.R"))
source(here::here("R/calc_sw_in.R"))

if (!dir.exists(rf_sample_data_tiles_dir)) {
  dir.create(rf_sample_data_tiles_dir, recursive = TRUE)
  message("Directory created: ", rf_sample_data_tiles_dir)
}
```

# 1. Overview

This script tests the data preprocessing pipeline on a single 1°×1° tile. The workflow includes:

1. Identifying valid windows (0.05° resolution) based on correlation analysis
2. Selecting valid 1°×1° tiles for sampling
3. Processing environmental layers for a single test tile
4. Creating spatial windows and sampling points
5. Calculating radiation index (Rin)

# 2. Input Data

## 2.1 Valid Windows (0.05° resolution)

After fixed window analysis, we have r(H,TWI) and r(H,Rin) at 0.05° resolution. 
Windows with < 30 valid observations (complete cases for H and TWI or H and Rin) are excluded.

```{r valid_win, eval=TRUE}
# Load correlation rasters
r_twi_vegh <- rast(cor_twi_vegh_mosaic_file)
r_h_rin <- rast(r_H_R_5km_path)

# Identify valid windows (both correlations available)
valid_win <- !is.na(r_twi_vegh) & !is.na(r_h_rin)

# Summary
cat("Valid windows summary:\n")
print(table(values(valid_win), useNA = "no"))

# Visualize
plot(valid_win, main = "Valid Windows (0.05°)")
```

## 2.2 Valid 1°×1° Tiles for Sampling

Aggregate valid windows to 1° resolution to identify tiles containing at least one valid window.

```{r valid_tile, eval=TRUE}
# Create 1° template
align_1_deg <- create_aligned_template(valid_win, res_out = 1)

# Aggregate to 1° tiles
# Only consider pixels with cover fraction > 0.1 to eliminate border effects
grid_1_deg <- raster_preprocess_save(
  input        = valid_win, 
  output       = NULL,
  target       = align_1_deg,
  varname      = "valid_tile",
  if_zonal     = TRUE,
  fun          = function(values, coverage_fractions) {
    valid_indices <- !is.na(values) & coverage_fractions > 0.1
    if (any(valid_indices)) {
      any(values[valid_indices] == 1, na.rm = TRUE)
    } else {
      FALSE
    }
  },
  if_aggregate = FALSE,
  if_resample  = FALSE,
  if_return_raster = TRUE
)

# Summary
cat("Valid tiles summary:\n")
print(table(values(grid_1_deg), useNA = "no"))

# Visualize
plot(grid_1_deg, main = "Valid 1°×1° Tiles")
```

## 2.3 Tile Coordinates

Convert raster to dataframe with tile boundaries.

```{r valid_tile_df, eval=FALSE}
rf_tiles <- as.data.frame(grid_1_deg, xy = TRUE) |>
  filter(lyr.1 == TRUE) |>
  transmute(
    xmin = floor(x - 0.5),
    xmax = ceiling(x + 0.5),
    ymin = floor(y - 0.5),
    ymax = ceiling(y + 0.5)
  ) |>
  mutate(
    reg_id = paste0(make_lat_label(ymax), "_", make_lon_label(xmin))
  )

cat("Number of valid tiles:", nrow(rf_tiles), "\n")
head(rf_tiles)

# "83N_019W" %in% rf_tiles$reg_id
```
# 3. Test Tile Processing

## 3.1 Select Test Tile


```{r select_tile}
rf_tiles <- arrow::read_parquet(rf_tiles_path)

# Select a specific tile for testing
reg_id <- "00N_025E"
# reg_id <- rf_tiles$reg_id[1]
cat("Testing tile", reg_id, "of", nrow(rf_tiles), "\n")

# Define tile extent
reg_extent <- terra::ext(
  rf_tiles$xmin[rf_tiles$reg_id==reg_id], 
  rf_tiles$xmax[rf_tiles$reg_id==reg_id], 
  rf_tiles$ymin[rf_tiles$reg_id==reg_id], 
  rf_tiles$ymax[rf_tiles$reg_id==reg_id]
)

# Create high-resolution valid window mask (0.00025° = 0.05°/200)
 valid_win_hr <- terra::crop(rast(valid_win_path), reg_extent) |>
      terra::disagg(fact = 200) 

# valid_win_hr <- terra::crop(valid_win, reg_extent) |> 
#  terra::disagg(fact = 200)

cat("Valid window mask created:\n")
print(valid_win_hr)
```

## 3.2 Visualize Tile Data

```{r visualize_tile}
# Plot correlation rasters for this tile
par(mfrow = c(1, 4))
plot(terra::crop(rast(cor_twi_vegh_mosaic_file), reg_extent), 
     main = "r(H,TWI) - Test Tile")
plot(terra::crop(rast(r_H_R_5km_path), reg_extent), 
     main = "r(H,Rin) - Test Tile")
plot(valid_win_hr, main = "Valid Windows")
plot(terra::crop(grid_1_deg, reg_extent), main = "Valid grid")
par(mfrow = c(1, 1))
```

## 3.3 Load Input Layers

```{r load_layers}
# Topographic Wetness Index
twi <- rast(twi_30m_path) |> 
  terra::crop(reg_extent)
twi <- twi / 100  # Scale to original values
cat("TWI range:", range(values(twi), na.rm = TRUE), "\n")

# Vegetation height
vegh <- extent_to_tile_ids(
  reg_extent, 
  tile_size = 3, 
  return_raster = TRUE, 
  source = "lang_vegh_10m", 
  tiles_dir = vegh_10m_tiles_dir
) |> terra::resample(twi, method = "bilinear")

# DEM
dem <- extent_to_tile_ids(
  reg_extent, 
  tile_size = 1, 
  return_raster = TRUE, 
  source = "copernicus_dem_30m", 
  tiles_dir = dem_30m_copernicus_dir
)

# Calculate slope and aspect, resample to TWI resolution
aligned <- aggregate_topography(
  dem,
  res_tar = NULL,
  target = twi,
  if_aggregate = FALSE,
  if_resample = TRUE
)

# Climate data
mat <- rast(mat_1km_file) |> 
  terra::crop(reg_extent) |> 
  terra::resample(twi, method = "bilinear")

map <- rast(map_1km_file) |> 
  terra::crop(reg_extent) |> 
  terra::resample(twi, method = "bilinear")

srad <- rast(srad_1km_file) |> 
  terra::crop(reg_extent) |> 
  terra::resample(twi, method = "bilinear")

# Visualize key layers
par(mfrow = c(3, 3))

plot(twi, main = "TWI")                    
plot(vegh, main = "Height (m)")
plot(aligned[["dem"]], main = "DEM (m)")   
plot(aligned[["slope"]], main = "Slope (deg)")  
plot(aligned[["aspect"]], main = "Aspect (deg)") 
plot(mat, main = "MAT (°C)")            
plot(map, main = "MAP (mm)")               
plot(srad, main = "Radiation (kJ m⁻² day⁻¹)") 

par(mfrow = c(1, 1))
```

# 4. Data Extraction and Sampling

## 4.1 Stack and Mask Layers

```{r stack_and_mask}
set.seed(124563)  # For reproducibility

# Stack all layers and apply valid window mask
stacked <- c(twi, vegh, aligned[["dem"]], aligned[["slope"]], 
             aligned[["aspect"]], mat, map, srad) |>
  terra::mask(valid_win_hr, maskvalues = FALSE)

cat("Stacked raster dimensions:", dim(stacked), "\n")
cat("Valid pixels after masking:", sum(!is.na(values(stacked[[1]]))), "\n")
```

## 4.2 Create Spatial Windows and Sample

```{r sample_data}
# Convert to dataframe, create windows, and sample
df_samp <- as.data.frame(stacked, xy = TRUE, na.rm = TRUE) |> 
  create_spatial_windows(
    coord_vars = c("lon", "lat"), 
    value_vars = c("twi", "vegh", "dem", "slope", "aspect", "mat", "map", "srad"), 
    dwin = 0.05
  ) |>
  tidyr::drop_na() |> 
  group_by(lon_mid, lat_mid) |>
  slice_sample(n = 5) |>
  ungroup()

cat("Total samples extracted:", nrow(df_samp), "\n")
cat("Number of unique windows:", n_distinct(df_samp$lon_mid, df_samp$lat_mid), "\n")
```

## 4.3 Calculate Radiation Index (Rin)

```{r calculate_rin}
# Calculate incoming solar radiation
df_samp <- df_samp |>
  mutate(
    sw_in_uneven = calc_sw_in(lat, slope, aspect, year = 2020),
    sw_in_flat = calc_sw_in(lat, 0, 0, year = 2020),
    rin = sw_in_uneven / sw_in_flat
  ) |>
  dplyr::select(lon, lat, vegh, dem, twi, rin, mat, map, srad)

cat("Radiation index (Rin) range:", range(df_samp$rin, na.rm = TRUE), "\n")

# Save data frame per tile
out_file <- file.path(rf_sample_data_tiles_dir, sprintf("tile_%s.parquet", reg_id))
arrow::write_parquet(df_samp, out_file)
if(file.exists(out_file)) message("✅ Saved: ", out_file)
```

# 5. Results Summary

```{r summary}
# Display first few rows
cat("Sample data (first 6 rows):\n")
print(head(df_samp))

# Summary statistics
cat("\nSummary statistics:\n")
summary(df_samp)

# Check for missing values
cat("\nMissing values per column:\n")
colSums(is.na(df_samp))
```

# 6. Next Steps

1. **Scale up**: Apply pipeline to all valid tiles
2. **Parallel processing**: Implement parallelization for multiple tiles and Write sampled data to parquet/CSV files
3. **Merge results**: Merge sampled data to parquet/CSV files
4. **RF modeling**: Use sampled data for random forest training



```{r, include=FALSE, eval=FALSE}
# This code will neither run nor show in the output
rmarkdown::render(here::here("vignettes/7_01_re1.Rmd"))
```
