---
title: "Aggregate GeoTIFF"
author: "Beni Stocker and Ting Tan"
date: "2025-02-19"
output: html_document
---

```{r setup, include=FALSE}
library(terra)
library(stringr)
library(here)
library(fs)
library(purrr)
library(furrr)
library(progress)
library(progressr)
library(cli) 
library(here)


source(here::here("R/aggregate_byfile.R"))
```

# Aggregation
The very high-resolution global canopy height dataset ([Lang et al., 2023](https://www.nature.com/articles/s41559-023-02206-6)) is available for download from [ETH Zurich](https://doi.org/10.3929/ethz-b-000609802). In contrast, the global Topographic Wetness Index (TWI) dataset ([Marthew, 2015](https://hess.copernicus.org/articles/19/91/2015/)) has a coarser resolution of 15 arcseconds (~450m at the Equator) and is available from the [CEH Catalogue](https://catalogue.ceh.ac.uk/documents/6b0c4358-2bf3-4924-aa8f-793d468b92be). To ensure compatibility for further analysis, the global canopy height data must first be aggregated to match the 15 arcsecond resolution of the TWI dataset.

```{r configuration, eval = TRUE}
# File paths
dir_vegh_450m <- file.path("/data_2/scratch/ting/vegh_450m/3_3_deg/")  # Path for saving modified data
dir_vegh_10m <- file.path("/data_2/archive/vegheight_lang_2023/data/3deg_cogs/")  # Path for higher resolution data
file_ga2 <- file.path("/data/archive/gti_marthews_2015/data/ga2.nc")  # File path for target raster
```

```{r agregation, eval = FALSE}
# Get .tif files from source directory
files_vegh_10m_all <- as.character(
  fs::dir_ls(path = dir_vegh_10m, glob = "*_Map.tif"))
files_vegh_10m <- files_vegh_10m_all[2211:2651]  # Select files for testing

# Set up parallel processing (4 workers, adjust based on CPU cores)
plan(multisession, workers = 4)

# Enable progress bar with `cli` style
handlers("cli")  

# Progress bar tracking
with_progress({
  pb <- progressor(along = files_vegh_10m)  # Create progress bar

  results <- future_map(files_vegh_10m, safely(~{
    raster_reference <- terra::rast(file_ga2)
    result <- aggregate_byfile(.x, raster_reference, dir_vegh_450m)  

    pb()  # Update progress bar at each step
    return(result)  # Return result
  }))
})

# Switch back to single-threaded execution after completion
plan(sequential)

# # Print results (optional)
# print(results) # here no info about results!!!!!!!!!

```

# Combination
After aggregating the global canopy height data to match the 15 arcsecond (~450m) resolution of the TWI dataset, the next step is to combine multiple raster tiles into a single mosaic for better 

```{r combinaitoin, eval = TRUE}
# Load processed rasters
files_vegh_450m <- as.character(
  fs::dir_ls(path = dir_vegh_450m, glob = "*_to450m.nc")) 
rasters_vegh_450m <- lapply(files_vegh_450m, terra::rast)  # Convert to SpatRaster objects

# Merge rasters into a single mosaic
raster_mosaic <- do.call(terra::merge, rasters_vegh_450m)

# define the path and name of the merged mosaic
file_mosaic <- file.path(dirname(dir_vegh_450m), "vegh_450m_2020_mosaic.nc")

# Save the merged raster as a NetCDF file
terra::writeCDF(raster_mosaic, file_mosaic, overwrite = TRUE)

# Plot the mosaic raster after writeCDF
plot(terra::rast(file_mosaic))

```

```{r PDF, eval = FALSE}
rmarkdown::render(here::here("vignettes", "aggregate_geotiff.Rmd"))
```
